% ============================================
% Spectrograms: A Mathematical and Computational Guide
% Educational manual for FFT-Based Audio and Image Processing
% ============================================

\section{Introduction}\label{sec:introduction}

This manual provides a comprehensive mathematical and computational
overview of the algorithms, optimizations, and transformations
implemented in the \texttt{spectrograms} library. The focus is on
understanding the theoretical foundations and practical
implementations of FFT-based signal processing for audio and image analysis.

The material emphasizes the \termdef{why} and \termdef{how} of each
algorithm: \emph{why} certain design choices matter, and \emph{how}
they are implemented efficiently. The material covers both the
mathematics and the practical computational considerations that make
these methods work in production systems.

\subsection{Target Audience and Prerequisites}\label{subsec:prerequisites}

This material assumes familiarity with:
\begin{itemize}[itemsep=3pt]
  \item Linear algebra (matrix multiplication, vector operations, eigenvalues)
  \item Complex numbers and Euler's formula
  \item Discrete signals and sampling theory
  \item Basic programming concepts and algorithm analysis
  \item Calculus (integrals, derivatives, series)
\end{itemize}

No prior knowledge of Fourier analysis, psychoacoustics, or advanced
signal processing is assumed.

\subsection{Scope and Organization}\label{subsec:organization}

The manual is organized into major thematic sections:
\begin{enumerate}[itemsep=3pt]
  \item \textbf{Fourier Transform Theory} (\cref{sec:fourier}) ---
    DFT, FFT algorithms, Hermitian symmetry, 2D FFT
  \item \textbf{Window Functions and Spectral Analysis}
    (\cref{sec:windows}) --- Hanning, Hamming, Kaiser, Blackman, Gaussian
  \item \textbf{Time-Frequency Representations} (\cref{sec:stft}) ---
    STFT and spectrograms
  \item \textbf{Amplitude Scaling and Representations} (\cref{sec:amplitude})
    --- Power, magnitude, decibels, Parseval's theorem
  \item \textbf{Perceptual Frequency Scales} (\cref{sec:perceptual})
    --- Mel, ERB, logarithmic scales
  \item \textbf{Constant-Q Transform} (\cref{sec:cqt}) --- Variable-length
    kernels, sparsity optimization
  \item \textbf{Advanced Audio Features} (\cref{sec:advanced}) ---
    MFCC, chroma, gammatone filters
  \item \textbf{Image Processing via FFT} (\cref{sec:image}) --- 2D
    convolution, filtering, edge detection
  \item \textbf{Binaural Spectrograms} (\cref{sec:binaural}) ---
    ITD, IPD, ILD, ILR spatial audio features
  \item \textbf{Modified Discrete Cosine Transform} (\cref{sec:mdct}) ---
    MDCT/IMDCT, perfect reconstruction, codec applications
  \item \textbf{FFT Convolution and Minimum Phase} (\cref{sec:convolution}) ---
    FFT-based convolution, streaming overlap-save, minimum-phase conversion
  \item \textbf{Computational Optimizations}
    (\cref{sec:optimizations}) --- Sparse matrices, FFT plans, caching
  \item \textbf{Numerical Considerations} (\cref{sec:numerical}) ---
    Precision, normalization, dynamic range
\end{enumerate}

\subsection{Numeric Precision: Generic over f32 and f64}\label{subsec:precision}

The library is \termdef{precision-generic}: each core computation is
written once and instantiated at either single (\code{f32}) or double
(\code{f64}) precision. This genericity is mediated by a single sealed
trait, \code{Sample}, which serves as the dispatch seam between the
precision-agnostic algorithms and the concrete floating-point type and
its FFT backend.

\begin{concept}{The Sample Trait as a Dispatch Seam}
  \code{Sample} is a \emph{sealed} trait, implemented only for
  \code{f32} and \code{f64}. It bundles the numeric capabilities the
  algorithms require (\code{Float}, \code{FloatConst}, \code{NumAssign},
  \code{Copy}, \code{Send}, \code{Sync}, and \code{'static}) together
  with the scalar conversions \code{from\_f64} and \code{from\_usize}
  (used to lift literal constants and index or length values into the
  chosen precision) and the FFT-plan constructors (\code{plan\_r2c},
    \code{plan\_c2r}, \code{plan\_c2c}, and their two-dimensional
  counterparts). The realfft and FFTW backend plan traits
  (\code{R2cPlan}, \code{C2cPlan}, \code{C2rPlan}, and their 2D
  variants) are themselves generic over the scalar type, so a single
  generic code path drives both precisions without duplication.
\end{concept}

\textbf{Default to f64 for backward compatibility.} The scalar type
parameter \emph{defaults} to \code{f64}. Consequently every \code{f64}
signature and example in this manual remains valid and unchanged in
meaning: double precision is selected automatically when the inputs are
\code{f64}, or when no type is specified. Single precision is strictly
opt-in---it is inferred from \code{f32} inputs, or requested explicitly
with turbofish syntax:

\begin{lstlisting}[caption={The same call at the default (f64) and at f32}]

// Double precision (default): T is inferred as f64
let s = stft(&sig_f64, n_fft, hop, WindowType::Hanning, true)?;

// Single precision: inferred from f32 inputs ...
let s = stft(&sig_f32, n_fft, hop, WindowType::Hanning, true)?;

// ... or requested explicitly via turbofish
let s = stft::<f32>(&sig_f32, n_fft, hop, WindowType::Hanning, true)?;
\end{lstlisting}

Precision genericity spans the full feature set: the STFT, the mel,
ERB, and CQT perceptual transforms, MFCCs, chroma, the MDCT, FFT
convolution, minimum-phase reconstruction, the binaural features, and
the 2D FFT and image operations all operate at both precisions. The
legacy \code{mdct\_f32} and \code{imdct\_f32} entry points have been
\emph{removed}; use the generic \code{mdct::<f32>} and
\code{imdct::<f32>} instead.

\begin{concept}{Caveat: f64-Returning Convenience Constructors}
  A few high-level convenience constructors that accept \code{f64}
  samples---\code{chromagram}, \code{mfcc}, and
  \code{gaussian\_kernel\_2d}---deliberately return \code{f64} results.
  They are ergonomic entry points for the common double-precision case
  and are \emph{not} generic. Single precision for these features is
  reached instead through the input-generic variants
  (\code{chromagram\_from\_spectrogram} and \code{mfcc\_from\_log\_mel})
  or through the low-level primitives, all of which are generic over
  the scalar type.
\end{concept}

\textbf{Motivation.} Supporting \code{f32} alongside \code{f64} serves
three concrete goals. First, many spectrogram workloads are
\emph{memory-bound}: halving the width of every sample and spectral
coefficient improves cache utilisation and memory bandwidth. Second,
machine-learning pipelines train in single precision, so producing
features directly in \code{f32} avoids a conversion at the
pipeline boundary. Third, modern hardware---both SIMD-capable CPUs and
GPUs---processes single-precision data at higher throughput, so
computing features in \code{f32} reduces wall-clock time for large-scale
batch workloads. Double precision nonetheless remains the default, as it
is the right choice for offline analysis where numerical headroom
matters (\cref{subsec:precision-f64}).

\subsection{Notation Conventions}\label{subsec:notation}

Throughout this manual:
\begin{itemize}[itemsep=2pt]
  \item $N$ denotes signal length or FFT size
  \item $k$ denotes frequency bin index ($0 \leq k < N$)
  \item $n$ denotes time/sample index ($0 \leq n < N$)
  \item $j = \sqrt{-1}$ (the imaginary unit)
  \item $X[k]$ denotes frequency domain values
  \item $x[n]$ denotes time domain values
  \item Vectors are bold: $\mathbf{x}$
  \item Complex conjugate: $\conj{z}$
\end{itemize}

\section{Fourier Transform Foundations}\label{sec:fourier}

\subsection{The Discrete Fourier Transform (DFT)}\label{subsec:dft}

The \termdef{Discrete Fourier Transform} (DFT) decomposes a
finite-length sequence into its constituent frequencies. For a
sequence $x[n]$ of length $N$, the DFT is defined
as~\cite{oppenheim1999discrete}:

\begin{equation}
  \label{eq:dft_definition}
  X[k] = \sum_{n=0}^{N-1} x[n] \cdot e^{-j2\pi kn/N}, \quad k = 0, 1,
  \ldots, N-1
\end{equation}

where $X[k]$ represents the complex amplitude at frequency bin $k$.
Using Euler's formula $e^{j\theta} = \cos\theta + j\sin\theta$,
\cref{eq:dft_definition} can be expanded as:

\begin{equation}
  \label{eq:dft_expanded}
  X[k] = \sum_{n=0}^{N-1} x[n] \left[\cos\left(\frac{2\pi
  kn}{N}\right) - j\sin\left(\frac{2\pi kn}{N}\right)\right]
\end{equation}

This reveals that the DFT correlates the input signal with complex
sinusoids at each frequency. The computational complexity of this
direct calculation is $\mathcal{O}(N^2)$ --- each of $N$ outputs
requires $N$ multiplications.

The \termdef{inverse DFT} (IDFT) reconstructs the time-domain signal:

\begin{equation}
  \label{eq:idft}
  x[n] = \frac{1}{N}\sum_{k=0}^{N-1} X[k] \cdot e^{j2\pi kn/N}, \quad
  n = 0, 1, \ldots, N-1
\end{equation}

The normalization factor $1/N$ ensures \termdef{perfect reconstruction}:
\[
  \IDFT(\DFT(x)) = x.
\]

\begin{concept}{DFT Scaling Convention}
  Different implementations use different scaling conventions for the
  DFT and inverse DFT.\ The three most common conventions are:

  \begin{enumerate}[itemsep=2pt]
    \item \textbf{No forward scaling (this library):} Forward DFT
      uses coefficient 1, inverse DFT uses $1/N$
    \item \textbf{Symmetric scaling:} Both forward and inverse use
      $1/\sqrt{N}$ (unitary transform)
    \item \textbf{No inverse scaling:} Forward uses $1/N$, inverse uses 1
  \end{enumerate}

  \textbf{This library uses Convention 1:} The forward DFT has
  \emph{no normalization factor} (coefficient of 1), and the inverse
  DFT includes the $1/N$ factor. This convention is standard in most
  signal processing literature and libraries (FFTW, NumPy, SciPy).

  \textbf{Energy implications:} With this scaling, Parseval's theorem
  (\cref{subsec:parseval}) takes the form:
  \begin{equation}
    \sum_{n=0}^{N-1} \abs{x[n]}^2 = \frac{1}{N}\sum_{k=0}^{N-1} \abs{X[k]}^2
  \end{equation}

  When comparing spectral amplitudes between implementations,
  \emph{always verify the scaling convention} to ensure
  energy-consistent analysis.
\end{concept}

\subsection{Fast Fourier Transform (FFT): The Cooley-Tukey
Algorithm}\label{subsec:fft}

The \termdef{Fast Fourier Transform} (FFT) is not a different
transform, but an efficient algorithm for computing the DFT.\ The
Cooley-Tukey algorithm~\cite{cooley1965algorithm}, published in 1965
(though similar ideas date to Gauss in
1805~\cite{gauss1866nachlass}), reduces complexity from
$\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$.

The key insight is divide-and-conquer using the \emph{symmetry} and
\emph{periodicity} properties of complex exponentials:

\begin{concept}{Twiddle Factor Properties}
  Let $W_N^{k} = e^{-j2\pi k/N}$ be the \termdef{twiddle factor}.
  These complex exponentials satisfy:
  \begin{align}
    \label{eq:twiddle_symmetry}
    W_N^{k+N/2} &= -W_N^{k} \quad \text{(symmetry)} \\
    \label{eq:twiddle_periodicity}
    W_N^{k+N} &= W_N^{k} \quad \text{(periodicity)}
  \end{align}

  \textbf{Periodicity} means the complex exponential repeats every
  $N$ samples. \textbf{Symmetry} means that values half an FFT bin
  apart differ only in sign. These properties enable the FFT's
  recursive decomposition.
\end{concept}

For $N$ even, split the DFT into even and odd indexed samples:

\begin{equation}
  \label{eq:fft_split}
  X[k] = \sum_{n=0}^{N/2-1} x[2n] W_N^{2nk} + \sum_{n=0}^{N/2-1}
  x[2n+1] W_N^{(2n+1)k}
\end{equation}

Using $W_N^{2k} = W_{N/2}^{k}$, this becomes:

\begin{equation}
  \label{eq:fft_butterfly}
  X[k] = \underbrace{\sum_{n=0}^{N/2-1} x[2n] W_{N/2}^{nk}}_{E[k]} +
  W_N^k \underbrace{\sum_{n=0}^{N/2-1} x[2n+1] W_{N/2}^{nk}}_{O[k]}
\end{equation}

where $E[k]$ and $O[k]$ are $N/2$-point DFTs of the even and odd
samples. This recursive splitting continues until the base cases are reached
(typically $N=1$ or small prime factors).

The complexity analysis: At each of $\log_2 N$ levels, $N$ operations
(additions and a twiddle factor multiplication) are performed, yielding
$\mathcal{O}(N \log N)$ total complexity.

\subsection{Real-to-Complex FFT and Hermitian Symmetry}\label{subsec:r2c_fft}

When the input signal $x[n]$ is real-valued (as in audio processing),
the DFT output exhibits a special property:

\begin{concept}{Hermitian Symmetry}
  For a real-valued signal $x[n]$, the DFT satisfies
  \termdef{Hermitian symmetry}:
  \begin{equation}\label{eq:hermitian_symmetry}
    X[k] = \conj{X}[N-k]
  \end{equation}

  where $\conj{X}$ denotes complex conjugate. This means:
  \begin{itemize}[itemsep=2pt]
    \item $\Real(X[k]) = \Real(X[N-k])$ (real parts are symmetric)
    \item $\Imag(X[k]) = -\Imag(X[N-k])$ (imaginary parts are antisymmetric)
    \item $X[0]$ and $X[N/2]$ (for even $N$) are purely real
  \end{itemize}

  This property follows directly from the DFT definition and the fact
  that $e^{j\theta} + e^{-j\theta} = 2\cos\theta$ is real for real $\theta$.

  \textbf{Critical constraint for R2C transforms:} For even $N$, the
  Nyquist frequency bin $X[N/2]$ \emph{must be purely real} (i.e.,
  $\Imag(X[N/2]) = 0$). This follows from Hermitian symmetry since
  $X[N/2] = \conj{X}[N - N/2] = \conj{X}[N/2]$, which requires the
  imaginary part to be zero. Violating this constraint results in
  non-real values after inverse FFT, causing numerical errors and
  incorrect signal reconstruction.
\end{concept}

This symmetry makes it sufficient to store only the positive frequencies:

\begin{equation}
  k \in \{0, 1, 2, \ldots, N/2\}
\end{equation}

This yields $N/2 + 1$ complex values instead of $N$, halving
storage requirements. The library computes real-to-complex (R2C) FFTs
with output size:

\begin{equation}
  \text{output\_size} = \lfloor N/2 \rfloor + 1
\end{equation}

\textbf{Implementation detail:} The library uses specialized R2C FFT
algorithms (from RealFFT or FFTW backends) that exploit this symmetry
for both speed and memory efficiency.

\begin{lstlisting}[caption={R2C output size for real FFT input}]

pub const fn r2c_output_size(n: usize) -> usize {
    n / 2 + 1
}

// Example: 512-point FFT on real signal
let samples = vec![0.0; 512];
let spectrum = fft(&samples, 512)?;
assert_eq!(spectrum.len(), 257); // 512/2 + 1
\end{lstlisting}

\subsection{Inverse FFT and Perfect Reconstruction}

The inverse FFT (IFFT) transforms frequency-domain data back to the
time domain. For a complex-to-real (C2R) inverse FFT, given
Hermitian-symmetric input $X[k]$:

\begin{equation}
  x[n] = \frac{1}{N}\sum_{k=0}^{N-1} X[k] e^{j2\pi kn/N}
\end{equation}

Due to Hermitian symmetry, it is sufficient to sum over the non-redundant
frequencies:

\begin{equation}
  x[n] = \frac{1}{N}\left[X[0] + 2\sum_{k=1}^{N/2-1}
  \text{Re}(X[k]e^{j2\pi kn/N}) + X[N/2]\cos(\pi n)\right]
\end{equation}

The library's C2R inverse FFT ensures perfect reconstruction:

\begin{lstlisting}[caption={Verifying C2R inverse FFT round-trip}]
let original = vec![1.0, 2.0, 3.0, 4.0];
let spectrum = fft2d(&original_image)?;
let reconstructed = ifft2d(&spectrum, original_image.ncols())?;
// reconstructed approximately original (within floating-point precision)
\end{lstlisting}

\subsection{2D FFT via Row-Column Decomposition}\label{subsec:fft2d}

For 2D data like images with dimensions $M \times N$, the 2D DFT is:

\begin{equation}
  \label{eq:dft_2d}
  X[k_1, k_2] = \sum_{n_1=0}^{M-1} \sum_{n_2=0}^{N-1} x[n_1, n_2]
  \cdot e^{-j2\pi(k_1 n_1/M + k_2 n_2/N)}
\end{equation}

The separability property allows this to be decomposed into row and
column transforms:

\begin{equation}
  \label{eq:fft_2d_decomposition}
  X[k_1, k_2] = \sum_{n_1=0}^{M-1} e^{-j2\pi k_1 n_1/M}
  \underbrace{\left[\sum_{n_2=0}^{N-1} x[n_1, n_2] \cdot e^{-j2\pi k_2
  n_2/N}\right]}_{\text{1D FFT along each row}}
\end{equation}

Algorithm:
\begin{enumerate}
  \item Apply 1D FFT to each row: $M$ transforms of length $N$
  \item Apply 1D FFT to each column of the result: $N$ transforms of length $M$
\end{enumerate}

For a real-valued $M \times N$ image, the output is $M \times (N/2 +
1)$ due to Hermitian symmetry along rows.

\textbf{Implementation:} The library uses row-major layout and
ensures contiguous memory:

\begin{lstlisting}[caption={2D FFT function signature and output shape}]

pub fn fft2d<T: Sample>(data: &ArrayView2<T>) -> SpectrogramResult<Array2<Complex<T>>> {
    let (nrows, ncols) = (data.nrows(), data.ncols());

    if !data.is_standard_layout() {
        return Err(SpectrogramError::invalid_input(
            "array must be contiguous and row-major"));
    }

    let out_shape = r2c_output_size_2d(nrows, ncols);
    // out_shape = (nrows, ncols/2 + 1)

    let mut plan = planner.plan_r2c_2d(nrows, ncols)?;
    // ... perform row-column FFT
}
\end{lstlisting}

Complexity: $O(MN \log N + MN \log M) = O(MN \log(MN))$ for square images.

\section{Window Functions for Spectral Analysis}\label{sec:windows}

\subsection{The Windowing Problem}\label{subsec:windowing_problem}

\begin{concept}{Spectral Leakage}
  When the DFT of a finite signal segment is computed, the signal is
  implicitly multiplied by a rectangular window (ones inside the
  segment, zeros outside). This abrupt truncation causes
  \termdef{spectral leakage}: energy from one frequency ``leaks''
  into adjacent frequency bins.

  Consider a pure sinusoid at frequency $f_0$. If sampling captures
  exactly an integer number of periods, the DFT produces a perfect spike at
  $f_0$. But with non-integer periods, the energy spreads across many bins.

  This occurs because multiplying by a rectangular window in the time
  domain is equivalent to convolving with a sinc function in the
  frequency domain, which has slowly-decaying sidelobes.
\end{concept}

The fundamental trade-off~\cite{harris1978use}:
\begin{itemize}[itemsep=3pt]
  \item \textbf{Narrow main lobe} $\rightarrow$ better frequency resolution
  \item \textbf{Low sidelobes} $\rightarrow$ less spectral leakage
\end{itemize}

Window functions smooth the transition to zero at the segment
boundaries, reducing leakage at the cost of frequency resolution.

\subsection{Windowing Convention: Symmetric
vs.\ Periodic}\label{subsec:windowing_convention}

\begin{concept}{Symmetric and Periodic Windows}
  Window functions can be defined with two distinct conventions:

  \textbf{Symmetric windows} use $N-1$ in the denominator:
  \begin{equation}
    w_{\text{sym}}[n] = f\left(\frac{n}{N-1}\right), \quad n \in [0, N-1]
  \end{equation}

  \textbf{Periodic windows} use $N$ in the denominator:
  \begin{equation}
    w_{\text{per}}[n] = f\left(\frac{n}{N}\right), \quad n \in [0, N-1]
  \end{equation}

  The difference is subtle but important:
  \begin{itemize}[itemsep=2pt]
    \item \textbf{Symmetric:} Values at endpoints match (e.g., both
      zero for Hanning). Standard for spectral analysis.
    \item \textbf{Periodic:} The window is periodic with period $N$,
      so $w[N] = w[0]$. Required for perfect reconstruction in
      overlap-add systems.
  \end{itemize}

  For the Constant Overlap-Add (COLA) property~\cite{harris1978use}:
  \begin{equation}
    \sum_{m=-\infty}^{\infty} w[n - mH] = C \quad \text{(constant)}
  \end{equation}
  where $H$ is the hop size. Periodic windows satisfy COLA for
  specific overlap ratios, enabling perfect signal reconstruction.

  \textbf{Library convention:} All window implementations in this
  library use the \emph{symmetric} formulation ($N-1$), which is
  standard for analysis applications. For synthesis requiring COLA,
  use appropriate overlap ratios (e.g., 50\% for Hanning).
\end{concept}

\subsection{Rectangular Window (No Windowing)}\label{subsec:rect_window}

The rectangular window is simply:
\begin{equation}
  \label{eq:rect_window}
  w[n] = 1, \quad n \in [0, N-1]
\end{equation}

Its frequency response has the narrowest main lobe but highest
sidelobes ($\approx$-13~dB), causing severe spectral leakage.

\textbf{Generic over precision.} The window constructors are generic
over the scalar type. The match-arm listings throughout this section
show the \code{f64} instantiation, whereas the public entry point has
the generic signature

\begin{lstlisting}[caption={Generic make-window constructor signature}]

pub fn make_window<T: Sample>(window: WindowType,
                              n_fft: NonZeroUsize) -> NonEmptyVec<T>
\end{lstlisting}

\noindent and the numeric literals (e.g.\ \code{0.5}, \code{0.54}) are
lifted to the chosen precision through \code{T::from\_f64} in the
generic code.

\begin{lstlisting}[caption={Rectangular window implementation}]

WindowType::Rectangular => {
    w.fill(1.0);
}
\end{lstlisting}

Use when: You need maximum frequency resolution and know the signal
contains exact integer periods.

\subsection{Hanning Window}\label{subsec:hann_window}

Also called the Hann window (named after Julius von Hann, not Richard
Hamming~\cite{harris1978use}), defined as:

\begin{equation}
  \label{eq:hann_window}
  w[n] = 0.5 - 0.5\cos\left(\frac{2\pi n}{N-1}\right), \quad n \in [0, N-1]
\end{equation}

This is a raised cosine function. It smoothly tapers to zero at both endpoints.

Properties:
\begin{itemize}[itemsep=2pt]
  \item Main lobe width: 4 bins (2$\times$ rectangular)
  \item First sidelobe: $\approx$-32~dB
  \item Sidelobe falloff: -18~dB/octave
\end{itemize}

\begin{lstlisting}[caption={Hanning window implementation}]

WindowType::Hanning => {
    let n1 = (n_fft - 1) as f64;
    for (n, v) in w.iter_mut().enumerate() {
        *v = 0.5 - 0.5 * cos(2.0 * PI * (n as f64) / n1);
    }
}
\end{lstlisting}

\textbf{Why this works:} The cosine taper reduces discontinuities at
the edges. In the frequency domain, the Hanning window's transform is
the convolution of the rectangular window's transform with two delta
functions, producing lower sidelobes.

The Hanning window is the most commonly used for general-purpose
spectral analysis.

\subsection{Hamming Window}\label{subsec:hamming_window}

Named after Richard Hamming~\cite{harris1978use}, this window uses
optimized coefficients:

\begin{equation}
  \label{eq:hamming_window}
  w[n] = 0.54 - 0.46\cos\left(\frac{2\pi n}{N-1}\right)
\end{equation}

The coefficients (0.54, 0.46) were chosen to minimize the first sidelobe:
\begin{itemize}
  \item Main lobe width: 4 bins (same as Hanning)
  \item First sidelobe: $\approx$-43 dB (better than Hanning)
  \item Sidelobe falloff: -6 dB/octave (worse than Hanning)
\end{itemize}

\begin{lstlisting}[caption={Hamming window implementation}]

WindowType::Hamming => {
    let n1 = (n_fft - 1) as f64;
    for (n, v) in w.iter_mut().enumerate() {
        *v = 0.54 - 0.46 * cos(2.0 * PI * (n as f64) / n1);
    }
}
\end{lstlisting}

Trade-off: The Hamming window doesn't reach exactly zero at the
endpoints (it's 0.08), which can cause issues for overlap-add
reconstruction. However, its lower first sidelobe makes it useful
when strong interfering signals exist.

\subsection{Blackman Window}\label{subsec:blackman_window}

A three-term cosine sum~\cite{blackman1958measurement} with excellent
sidelobe suppression:

\begin{equation}
  \label{eq:blackman_window}
  w[n] = 0.42 - 0.5\cos\left(\frac{2\pi n}{N-1}\right) +
  0.08\cos\left(\frac{4\pi n}{N-1}\right)
\end{equation}

Properties:
\begin{itemize}
  \item Main lobe width: 6 bins (3$\times$ rectangular)
  \item First sidelobe: $\approx$-58 dB
  \item Sidelobe falloff: -18 dB/octave
\end{itemize}

\begin{lstlisting}[caption={Blackman window implementation}]

WindowType::Blackman => {
    let n1 = (n_fft - 1) as f64;
    for (n, v) in w.iter_mut().enumerate() {
        let a = 2.0 * PI * (n as f64) / n1;
        *v = 0.42 - 0.5 * cos(a) + 0.08 * cos(2.0 * a);
    }
}
\end{lstlisting}

Use when: You need minimal spectral leakage and can tolerate wider
main lobes (reduced frequency resolution).

\subsection{Kaiser Window and Bessel Functions}\label{subsec:kaiser_window}

The Kaiser window~\cite{kaiser1974nonrecursive} provides a tunable
parameter $\beta$ that controls the trade-off between main lobe width
and sidelobe level:

\begin{equation}
  \label{eq:kaiser_window}
  w[n] =
  \frac{I_0\left(\beta\sqrt{1-{\left(\frac{2n}{N-1}-1\right)}^{2}}\right)}{I_0(\beta)}
\end{equation}

\begin{concept}{Modified Bessel Function}
  The \termdef{modified Bessel function of the first kind}, order
  zero, $I_0(x)$ is defined by:
  \begin{equation}
    \label{eq:bessel_i0}
    I_0(x) = \sum_{k=0}^{\infty} \frac{1}{(k!)^2}{\left(\frac{x}{2}\right)^{2k}}
  \end{equation}

  For computational efficiency, a truncated power series is used. The
  series converges rapidly; typically 20--30 terms achieve
  machine-precision accuracy for $x < 10$.
\end{concept}

Parameter $\beta$ effects:
\begin{itemize}
  \item $\beta = 0$: rectangular window
  \item $\beta = 5$: similar to Hamming
  \item $\beta = 8.6$: similar to Blackman
  \item Larger $\beta$: lower sidelobes, wider main lobe
\end{itemize}

\begin{lstlisting}[caption={Kaiser window with modified Bessel function}]

fn bessel_i0(x: f64) -> f64 {
    let mut sum = 1.0;       // k=0 term
    let mut term = 1.0;
    let half_x = x / 2.0;

    // Truncated series: sufficient for Kaiser windows
    for k in 1..30 {
        term *= (half_x / k as f64).powi(2);
        sum += term;

        // Early termination if converged
        if term < 1e-12 * sum {
            break;
        }
    }

    sum
}

WindowType::Kaiser { beta } => {
    let denominator = bessel_i0(beta);

    for i in 0..n_fft {
        let n = i as f64;
        let n_max = (n_fft - 1) as f64;
        let alpha = (n - n_max / 2.0) / (n_max / 2.0);
        let bessel_arg = beta * (1.0 - alpha * alpha).sqrt();

        w[i] = bessel_i0(bessel_arg) / denominator;
    }
}
\end{lstlisting}

The Kaiser window is named after James Kaiser who introduced it in
1966 for optimal FIR filter design.

\subsection{Gaussian Window}\label{subsec:gaussian_window}

A Gaussian function centered at the window midpoint:

\begin{equation}
  w[n] = \exp\left(-\frac{1}{2}\left(\frac{n -
  \frac{N-1}{2}}{\sigma}\right)^{2}\right)
\end{equation}

where $\sigma$ (standard deviation) controls the width. Smaller
$\sigma$ → narrower window → more time localization.

\begin{lstlisting}[caption={Gaussian window implementation}]

WindowType::Gaussian { std } => {
    for i in 0..n_fft {
        let n = i as f64;
        let center = (n_fft - 1) as f64 / 2.0;
        let exponent = -0.5 * ((n - center) / std).powi(2);
        w[i] = exp(exponent);
    }
}
\end{lstlisting}

The Gaussian window has a unique property: its Fourier transform is
also Gaussian. This gives optimal time-frequency localization per the
uncertainty principle:

\begin{equation}
  \Delta t \cdot \Delta f \geq \frac{1}{4\pi}
\end{equation}

Use for: Time-frequency analysis where minimizing uncertainty is
critical (e.g., Gabor transforms).

\subsection{Trade-offs: Frequency Resolution vs. Spectral Leakage}

Comparison table for $N=512$:

\begin{center}
  \begin{tabular}{lccc}
    \textbf{Window} & \textbf{Main Lobe} & \textbf{Peak Sidelobe} &
    \textbf{Falloff Rate} \\
    \hline
    Rectangular & 2 bins & -13 dB & -6 dB/oct \\
    Hanning & 4 bins & -32 dB & -18 dB/oct \\
    Hamming & 4 bins & -43 dB & -6 dB/oct \\
    Blackman & 6 bins & -58 dB & -18 dB/oct \\
    Kaiser ($\beta=8.6$) & 6 bins & -60 dB & -6 dB/oct \\
  \end{tabular}
\end{center}

\textbf{Selection guidelines:}
\begin{itemize}
  \item General purpose, good balance: Hanning
  \item Need low first sidelobe: Hamming
  \item Need very low sidelobes: Blackman
  \item Custom trade-off: Kaiser with tuned $\beta$
  \item Optimal uncertainty: Gaussian with appropriate $\sigma$

\end{itemize}

\subsection{Constant Overlap-Add (COLA) Constraint}\label{subsec:cola}

\begin{concept}{Perfect Reconstruction and COLA}
  When using the STFT for signal modification or synthesis (e.g.,
  time-stretching, pitch-shifting, noise reduction), the time-domain
  signal must be reconstructed from modified spectral frames.
  The \termdef{Constant Overlap-Add} (COLA) constraint ensures
  perfect reconstruction is possible~\cite{harris1978use}.

  For a window function $w[n]$ and hop size $H$, the COLA property
  states that the sum of overlapping windows must be constant:

  \begin{equation}
    \label{eq:cola}
    \sum_{m=-\infty}^{\infty} w[n - mH] = C \quad \text{(constant for
    all } n\text{)}
  \end{equation}

  where $C$ is a non-zero constant (typically normalized to 1).

  \textbf{Physical interpretation:} When frames are overlapped by
  hop size $H$ and summed (overlap-add), every sample in the
  reconstructed signal receives equal weighting from the windows.
  This prevents amplitude modulation artifacts.
\end{concept}

\textbf{COLA conditions for common windows:}

\begin{enumerate}[itemsep=3pt]
  \item \textbf{Rectangular window:} COLA for any hop size $H \leq N$
    (no overlap required)

  \item \textbf{Hanning window (periodic):} COLA for $H = N/2, N/4,
    N/8, \ldots$ (50\%, 75\%, 87.5\% overlap, etc.)
    \begin{itemize}
      \item 50\% overlap ($H = N/2$): Most common, $C = 1$
      \item 75\% overlap ($H = N/4$): Higher redundancy, smoother synthesis
    \end{itemize}

  \item \textbf{Hamming window:} COLA for $H \approx 0.375N$ to
    $0.5N$ (requires careful tuning; 50\% overlap gives $C \approx 1.08$)

  \item \textbf{Blackman window:} COLA for $H \leq N/3$ (at least
    67\% overlap required)

  \item \textbf{Kaiser window:} COLA depends on $\beta$; generally
    requires $50-75\%$ overlap
\end{enumerate}

\textbf{Verification:} To check if a window/hop combination satisfies
COLA, compute:

\begin{lstlisting}[caption={Verifying COLA constraint for window and hop}]

fn verify_cola(window: &[f64], hop_size: usize) -> bool {
    let n = window.len();
    let n_overlaps = (n + hop_size - 1) / hop_size;

    // Check multiple offset positions
    for offset in 0..hop_size {
        let mut sum = 0.0;

        for m in 0..n_overlaps {
            let idx = offset + m * hop_size;
            if idx < n {
                sum += window[idx];
            }
        }

        // All positions should have same sum (within tolerance)
        if (sum - window[0]).abs() > 1e-10 {
            return false;
        }
    }

    true
}
\end{lstlisting}

\textbf{Implications for spectrogram inversion:}

When using the STFT for analysis-modification-synthesis:
\begin{enumerate}
  \item Choose a window/hop combination that satisfies COLA
  \item Apply STFT to get frames $X[m, k]$
  \item Modify the spectrogram (e.g., filter, denoise, time-stretch)
  \item Apply inverse FFT to each modified frame
  \item Overlap-add the frames with hop size $H$
  \item Divide by the COLA constant $C$ (if $C \neq 1$)
\end{enumerate}

The reconstructed signal $\hat{x}[n]$ will equal the original $x[n]$
if no modifications were made (perfect reconstruction).

\textbf{Library convention:} For analysis-only applications
(spectrograms, feature extraction), COLA is not required. However,
this library uses 50\% overlap ($H = N/2$) by default, which
satisfies COLA for Hanning windows and enables optional synthesis workflows.

\section{Short-Time Fourier Transform (STFT)}\label{sec:stft}

\subsection{Time-Frequency Analysis Motivation}\label{subsec:stft_motivation}

The standard DFT (\cref{eq:dft_definition}) provides frequency
content but discards all temporal information: it indicates \emph{which}
frequencies are present, but not \emph{when} they occur. For
non-stationary signals (speech, music, transient events),
time-frequency analysis is required~\cite{allen2003short}.

The \termdef{Short-Time Fourier Transform} (STFT) solves this by
applying the DFT to short, overlapping segments (frames) of the
signal, creating a time-frequency representation.

\subsection{STFT Definition and Computation}\label{subsec:stft_definition}

For a signal $x[n]$ with window function $w[n]$ of length $N$, the STFT is:

\begin{equation}
  \label{eq:stft}
  X[m, k] = \sum_{n=0}^{N-1} x[n + mH] \cdot w[n] \cdot e^{-j2\pi kn/N}
\end{equation}

where:
\begin{itemize}[itemsep=2pt]
  \item $m$ is the frame index
  \item $H$ is the \termdef{hop size} (samples between consecutive frames)
  \item $k$ is the frequency bin
  \item $N$ is the FFT size (window length)
\end{itemize}

The result is a 2D complex matrix: $X[m, k]$ with dimensions (time
frames) $\times$ (frequency bins).

\subsection{Framing and Overlap}

The signal is divided into overlapping frames:
\begin{equation}
  \text{Frame } m \text{ starts at sample: } mH
\end{equation}

Key parameters:
\begin{itemize}
  \item \textbf{FFT size ($N$):} Window length, determines frequency
    resolution: $\Delta f = f_s / N$
  \item \textbf{Hop size ($H$):} Samples between frames, determines
    time resolution: $\Delta t = H / f_s$
  \item \textbf{Overlap:} $(N - H) / N$. Common: 50\% (H = N/2) or
    75\% (H = N/4)
\end{itemize}

Number of frames for signal of length $L$ samples:
\begin{equation}
  M = \left\lfloor\frac{L - N}{H}\right\rfloor + 1
\end{equation}

With centering (padding $N/2$ zeros on each end):
\begin{equation}
  M = \left\lfloor\frac{L + N - H}{H}\right\rfloor
\end{equation}

\begin{concept}{STFT Centering and Temporal Alignment}
  \textbf{Why pad $N/2$ zeros?} Without centering, the first window
  is positioned with its \emph{left edge} at sample 0. This means the
  window's center is at sample $N/2$, causing a temporal shift in the
  resulting spectrogram.

  By padding $N/2$ zeros \emph{before} the signal, the first window's
  \emph{center} aligns with sample 0. This ensures:
  \begin{itemize}[itemsep=2pt]
    \item Frame $m$ is centered at time $t = mH/f_s$ (no offset)
    \item Edge effects are symmetric at signal boundaries
    \item Improved temporal localization for transient events
  \end{itemize}

  The same $N/2$ padding is applied at the signal's end, ensuring
  symmetric treatment of boundaries. This is standard practice in
  audio analysis libraries (e.g., librosa's \texttt{center=True}).

  \textbf{Padding type specification:} This library uses
  \emph{zero-padding} (also called zero-extension) for both the
  initial and final $N/2$ samples:
  \begin{equation}
    \tilde{x}[n] =
    \begin{cases}
      0 & n < 0 \\
      x[n] & 0 \leq n < L \\
      0 & n \geq L
    \end{cases}
  \end{equation}

  where $L$ is the original signal length.

  \textbf{Alternative: Reflective padding} extends the signal by
  mirroring boundary values (e.g., $\tilde{x}[-1] = x[0]$,
  $\tilde{x}[-2] = x[1]$), which can reduce edge discontinuities for
  certain signals. However, zero-padding is the standard convention
  in most audio processing libraries (NumPy, SciPy, librosa) and is
  used exclusively in this implementation.

  \textbf{Implementation consistency:} All implementations must use
  the same padding type to ensure identical spectrograms for the same
  input. Using reflective padding when zero-padding is expected (or
  vice versa) will cause boundary frames to diverge, particularly for
  the first and last $\lceil N/(2H) \rceil$ frames where padding
  contributes to the windowed signal.
\end{concept}

\textbf{Implementation detail:} The library uses centering by default
to avoid edge effects:

\begin{lstlisting}[caption={Frame count with optional centre-padding}]

fn frame_count(&self, n_samples: usize) -> usize {
    let pad = if self.centre { self.n_fft / 2 } else { 0 };
    let padded_len = n_samples + 2 * pad;

    if padded_len < self.n_fft {
        return 1;  // Single frame, even for short signals
    }

    let remaining = padded_len - self.n_fft;
    let n_frames = remaining / self.hop + 1;
    n_frames
}
\end{lstlisting}

\subsection{STFT Computation Pipeline}

For each frame $m$:

\begin{enumerate}
  \item \textbf{Extract frame:} Get samples $x[mH], x[mH+1], \ldots, x[mH+N-1]$
  \item \textbf{Apply window:} Compute $x_w[n] = x[mH + n] \cdot w[n]$
  \item \textbf{FFT:} Compute $X_m[k] = \text{FFT}(x_w)$
  \item \textbf{Store:} Save as column $m$ of output matrix
\end{enumerate}

Optimizations:
\begin{itemize}
  \item Pre-compute window samples once
  \item Reuse FFT plan across all frames
  \item Allocate workspace buffers once
\end{itemize}

\begin{concept}{Workspace Memory Requirements}
  To enable efficient STFT computation, implementations use a
  \termdef{Workspace} structure that holds reusable buffers and FFT
  state. This avoids repeated allocations during frame processing.

  \textbf{Minimum workspace requirements:}
  \begin{enumerate}[itemsep=2pt]
    \item \textbf{Windowed frame buffer:} Real-valued array of size
      $N$ to hold $x[mH + n] \cdot w[n]$ before FFT
    \item \textbf{FFT output buffer:} Complex-valued array of size
      $N/2 + 1$ (for R2C FFT) to hold frequency-domain values $X_m[k]$
    \item \textbf{FFT plan state:} Backend-specific data structure
      (e.g., FFTW plan, RealFFT planner) that stores twiddle factors
      and algorithm configuration
    \item \textbf{Power/magnitude spectrum buffer:} Real-valued array
      of size $N/2 + 1$ to hold $|X_m[k]|^2$ or $|X_m[k]|$ for the
      current frame
  \end{enumerate}

  Total memory requirement: $O(N)$ real values and $O(N)$ complex
  values, plus backend-specific FFT plan overhead (typically $O(N
  \log N)$ for precomputed twiddle factors).

  \textbf{Implementation example:}

  \begin{lstlisting}[caption={Workspace struct with reusable STFT buffers}]
struct Workspace {
    frame: Vec<f64>,              // Size N (windowed input)
    fft_out: Vec<Complex<f64>>,   // Size N/2 + 1 (FFT output)
    spectrum: Vec<f64>,           // Size N/2 + 1 (power/magnitude)
}

impl Workspace {
    fn new(n_fft: usize) -> Self {
        let out_len = n_fft / 2 + 1;
        Self {
            frame: vec![0.0; n_fft],
            fft_out: vec![Complex::zero(); out_len],
            spectrum: vec![0.0; out_len],
        }
    }
}
  \end{lstlisting}

  \textbf{Performance benefit:} Reusing workspace buffers across
  frames reduces memory allocator pressure. For a 10-second audio
  file at 16 kHz with $N=512$ and $H=256$, this processes
  approximately 625 frames --- avoiding 1,875 allocations (3 buffers
  $\times$ 625 frames).
\end{concept}

\begin{lstlisting}[caption={Windowed frame extraction and FFT per frame}]
 - compute one frame
fn compute_frame_spectrum(&mut self, samples: &[f64],
                          frame_idx: usize,
                          workspace: &mut Workspace) -> Result<()> {
    let out = workspace.frame.as_mut_slice();
    let pad = if self.centre { self.n_fft / 2 } else { 0 };
    let start = frame_idx * self.hop;

    // Fill windowed frame (with zero-padding)
    for i in 0..self.n_fft {
        let v_idx = start + i;
        let s_idx = v_idx as isize - pad as isize;

        let sample = if s_idx < 0 || s_idx >= samples.len() {
            0.0  // Padding
        } else {
            samples[s_idx as usize]
        };

        out[i] = sample * self.window[i];
    }

    // Compute FFT
    self.fft.process(out, workspace.fft_out)?;

    // Convert to power spectrum: |X[k]|^2
    for (i, c) in workspace.fft_out.iter().enumerate() {
        workspace.spectrum[i] = c.norm_sqr();
    }

    Ok(())
}
\end{lstlisting}

\subsection{Spectrogram as STFT Magnitude}

A spectrogram is the magnitude (or power) of the STFT:\

\begin{equation}
  S[m, k] = |X[m, k]|^2 \quad \text{(power spectrogram)}
\end{equation}

or

\begin{equation}
  S[m, k] = |X[m, k]| \quad \text{(magnitude spectrogram)}
\end{equation}

This discards phase information but provides an intuitive
visualization of energy distribution over time and frequency.

\subsection{Time-Frequency Uncertainty Principle}\label{subsec:uncertainty}

\begin{concept}{Time-Frequency Uncertainty Principle}
  The fundamental trade-off in STFT is governed by the uncertainty
  principle~\cite{gabor1946theory}:

  \begin{equation}
    \label{eq:uncertainty_principle}
    \Delta t \cdot \Delta f \geq \frac{1}{4\pi}
  \end{equation}

  where $\Delta t$ is temporal resolution and $\Delta f$ is frequency
  resolution.

  This is \emph{not} a limitation of the algorithm, but a fundamental
  property of Fourier analysis related to the Heisenberg uncertainty
  principle in quantum mechanics. You \textbf{cannot} achieve
  arbitrary precision in both domains simultaneously.
\end{concept}

Implications:
\begin{itemize}[itemsep=3pt]
  \item \textbf{Large $N$ (long window):} Good frequency resolution,
    poor time resolution
  \item \textbf{Small $N$ (short window):} Good time resolution, poor
    frequency resolution
\end{itemize}

For speech (16 kHz sample rate):
\begin{itemize}
  \item $N = 512$ (32 ms): $\Delta f = 31.25$ Hz --- good for rapid changes
  \item $N = 2048$ (128 ms): $\Delta f = 7.8$ Hz --- better frequency detail
\end{itemize}

For music (44.1 kHz sample rate):
\begin{itemize}
  \item $N = 2048$ (46 ms): $\Delta f = 21.5$ Hz --- standard choice
  \item $N = 4096$ (93 ms): $\Delta f = 10.8$ Hz --- for harmonic analysis
\end{itemize}

\textbf{Cannot achieve arbitrary precision in both domains
simultaneously.} This is not a limitation of the algorithm, but a
fundamental property of Fourier analysis related to the Heisenberg
uncertainty principle in quantum mechanics.

\section{Amplitude Scaling and Representations}\label{sec:amplitude}

\subsection{Power Spectrum}\label{subsec:power_spectrum}

The power spectrum represents energy at each frequency:

\begin{equation}
  P[k] = |X[k]|^2 = \text{Re}(X[k])^{2} + \text{Im}(X[k])^{2}
\end{equation}

For real input signals, this is proportional to the energy in frequency bin $k$.

Properties:
\begin{itemize}
  \item Always non-negative
  \item Units: (amplitude)$^2$
  \item Additive: energies from independent sources sum
\end{itemize}

The squared magnitude emphasizes strong components and suppresses
weak ones, which can be both beneficial (noise reduction) and
problematic (loss of weak signals).

\subsection{Magnitude Spectrum}

The magnitude spectrum is the absolute value:

\begin{equation}
  M[k] = |X[k]| = \sqrt{\text{Re}(X[k])^2 + \text{Im}(X[k])^{2}}
\end{equation}

This is more perceptually linear than power, as human perception of
sound intensity is approximately logarithmic.

Relationship:
\begin{equation}
  M[k] = \sqrt{P[k]}
\end{equation}

Use magnitude when:
\begin{itemize}
  \item You need perceptually meaningful amplitudes
  \item Comparing with other magnitude-based representations
  \item Visualization where extreme values would dominate
\end{itemize}

\subsection{Decibel Scale}\label{subsec:decibel}

The decibel (dB) scale provides logarithmic amplitude scaling:

\begin{equation}
  \label{eq:db_scale_ref}
  S_{\text{dB}}[k] = 10 \log_{10}\left(\frac{P[k]}{P_{\text{ref}}}\right)
\end{equation}

where $P_{\text{ref}}$ is a reference power. For power spectrograms
without specific reference:

\begin{equation}
  \label{eq:db_scale}
  S_{\text{dB}}[k] = 10 \log_{10}(P[k])
\end{equation}

For magnitude-based signals:

\begin{equation}
  \label{eq:db_scale_magnitude}
  S_{\text{dB}}[k] = 20 \log_{10}(M[k])
\end{equation}

The factor of 20 (instead of 10) accounts for the squared
relationship between magnitude and power.

\textbf{Numerical considerations:} To avoid $\log(0) = -\infty$, a
floor is applied:

\begin{equation}
  S_{\text{dB}}[k] = 10 \log_{10}(\max(P[k], \epsilon))
\end{equation}

where $\epsilon$ is a small threshold (e.g., $10^{-10}$). This maps
very small values to a large negative dB value instead of $-\infty$.

\textbf{Dynamic range compression:} The dB scale compresses the
dynamic range, making it easier to visualize spectrograms with both
loud and quiet components. A signal with power ratios of 1,000,000:1
becomes 60 dB, much more manageable.

Example:
\begin{itemize}
  \item Power = 1.0 → 0 dB
  \item Power = 0.1 → -10 dB
  \item Power = 0.01 → -20 dB
  \item Power = 100.0 → 20 dB
\end{itemize}

\subsection{Parseval's Theorem and Energy Conservation}\label{subsec:parseval}

\begin{concept}{Parseval's Theorem}
  Parseval's theorem states that energy is conserved between time and
  frequency domains:

  \begin{equation}
    \label{eq:parseval}
    \sum_{n=0}^{N-1} \abs{x[n]}^2 = \frac{1}{N}\sum_{k=0}^{N-1} \abs{X[k]}^2
  \end{equation}

  This fundamental result ensures that:
  \begin{itemize}[itemsep=2pt]
    \item The FFT doesn't create or destroy energy
    \item Total power in frequency domain equals total power in time domain
    \item Inverse FFT perfectly reconstructs energy (up to numerical precision)
  \end{itemize}
\end{concept}

For windowed signals in STFT with proper overlap-add, energy is preserved:

\begin{equation}
  \sum_m \sum_k |X[m,k]|^2 \propto \sum_n |x[n]|^2
\end{equation}

where the proportionality constant depends on the window normalization.

\textbf{Implication for normalization:} Different FFT libraries use
different normalization conventions. The library uses the convention
where the forward FFT has no normalization factor, and the inverse
FFT divides by $N$. This is consistent with most signal processing literature.

\section{Perceptual Frequency Scales}\label{sec:perceptual}

\subsection{Motivation: Human Auditory
Perception}\label{subsec:perceptual_motivation}

Auditory frequency perception is not linear~\cite{fletcher1940auditory}.
Frequency differences are perceived logarithmically, especially at
higher frequencies. For example:
\begin{itemize}[itemsep=2pt]
  \item The difference between 100~Hz and 200~Hz is highly
    perceptible (one octave)
  \item The difference between 10,000~Hz and 10,100~Hz is barely
    noticeable (0.014 octaves)
\end{itemize}

Perceptual frequency scales map linear Hz to
psychoacoustically-motivated scales that better match human hearing.

\subsection{Mel Scale}\label{subsec:mel_scale}

The mel scale is based on pitch perception
studies~\cite{stevens1937scale}. The name comes from ``melody,''
reflecting its musical origins. The conversion formulas are:

\begin{equation}
  \label{eq:hz_to_mel}
  \text{mel}(f) = 2595 \log_{10}\left(1 + \frac{f}{700}\right)
\end{equation}

\begin{equation}
  \label{eq:mel_to_hz}
  f(\text{mel}) = 700\left(10^{\text{mel}/2595} - 1\right)
\end{equation}

The mel scale is approximately linear below 1000~Hz and logarithmic above.

\textbf{Mel Filterbank Construction:}

To create $M$ mel-spaced filters:

\begin{enumerate}
  \item Convert frequency range $[f_{\min}, f_{\max}]$ to mel scale
  \item Create $M + 2$ equally-spaced mel points (for triangular filter edges)
  \item Convert mel points back to Hz
  \item Map Hz to FFT bin indices
  \item Create triangular filters centered at these points
\end{enumerate}

\begin{equation}
  H_m[k] =
  \begin{cases}
    0 & k < f[m-1] \\
    \frac{k - f[m-1]}{f[m] - f[m-1]} & f[m-1] \leq k < f[m] \\
    \frac{f[m+1] - k}{f[m+1] - f[m]} & f[m] \leq k \leq f[m+1] \\
    0 & k > f[m+1]
  \end{cases}
\end{equation}

where $f[m]$ are the FFT bin indices corresponding to mel points.

\textbf{Implementation detail:} The filterbank is stored as a sparse
matrix since each triangular filter only has non-zero values over a
small range of bins:

\begin{lstlisting}[caption={Building the sparse mel filterbank matrix}]

fn build_mel_filterbank_matrix(sample_rate_hz: f64, n_fft: usize,
                                n_mels: usize, f_min: f64,
                                f_max: f64) -> SparseMatrix {
    let out_len = r2c_output_size(n_fft);
    let df = sample_rate_hz / n_fft as f64;

    // Convert to mel scale
    let mel_min = hz_to_mel(f_min);
    let mel_max = hz_to_mel(f_max);

    // Create n_mels + 2 mel points (for triangle edges)
    let n_points = n_mels + 2;
    let step = (mel_max - mel_min) / (n_points - 1) as f64;

    let mel_points: Vec<_> = (0..n_points)
        .map(|i| i as f64 * step + mel_min)
        .collect();

    // Convert back to Hz, then to FFT bins
    let bin_points: Vec<_> = mel_points
        .iter()
        .map(|&mel| mel_to_hz(mel))
        .map(|hz| (hz / df).floor() as usize)
        .collect();

    // Build sparse filterbank
    let mut fb = SparseMatrix::new(n_mels, out_len);

    for m in 0..n_mels {
        let left = bin_points[m];
        let centre = bin_points[m + 1];
        let right = bin_points[m + 2];

        // Rising slope: left -> centre
        for k in left..centre {
            let v = (k - left) as f64 / (centre - left) as f64;
            fb.set(m, k, v);
        }

        // Falling slope: centre -> right
        for k in centre..right {
            let v = (right - k) as f64 / (right - centre) as f64;
            fb.set(m, k, v);
        }
    }

    fb
}
\end{lstlisting}

For typical speech processing ($N=512$, 80 mel bands), the filterbank
is approximately 95\% sparse.

\textbf{Numerical hygiene:} To achieve this sparsity in practice,
coefficients below a threshold (typically $10^{-12}$ or machine
  epsilon $\epsilon_{\text{machine}} \approx 2.2 \times 10^{-16}$ for
f64) are set to exactly zero during construction. This prevents
accumulation of negligible numerical artifacts while maintaining
filterbank fidelity.

\subsection{ERB Scale: Glasberg and Moore Model}\label{subsec:erb_scale}

The \termdef{Equivalent Rectangular Bandwidth} (ERB) scale is based
on psychoacoustic measurements of critical
bandwidths~\cite{glasberg1990derivation,moore1983suggested}. It
models the frequency selectivity of the human auditory system.

The ERB in Hz as a function of center frequency:

\begin{equation}
  \label{eq:erb_bandwidth}
  \text{ERB}(f) = 24.7 \left(4.37 \frac{f}{1000} + 1\right)
\end{equation}

To convert frequency to ERB scale (ERB-rate):

\begin{equation}
  \label{eq:hz_to_erb}
  \text{ERB-rate}(f) = 21.4 \log_{10}(4.37 \frac{f}{1000} + 1)
\end{equation}

Inverse transformation:

\begin{equation}
  \label{eq:erb_to_hz}
  f(\text{ERB}) = \frac{1000}{4.37}\left(10^{\text{ERB}/21.4} - 1\right)
\end{equation}

\textbf{Gammatone Filters:}

ERB spectrograms often use gammatone filters, which model the impulse
response of the auditory filter:

\begin{equation}
  g(t) = t^{n-1} e^{-2\pi b t} \cos(2\pi f_c t + \phi)
\end{equation}

where:
\begin{itemize}
  \item $n$ is the filter order (typically 4)
  \item $b$ is the bandwidth parameter related to ERB
  \item $f_c$ is the center frequency
  \item $\phi$ is the phase
\end{itemize}

The library implements gammatone filters in the frequency domain for efficiency.

\textbf{Why ERB over Mel?} The ERB scale more accurately models
auditory filters at higher frequencies and is preferred for
applications requiring high fidelity to human perception.

\subsection{Logarithmic Frequency (Musical Scale)}\label{subsec:log_frequency}

Musical notes are logarithmically spaced: each octave doubles the
frequency. For music analysis, a logarithmic frequency axis is natural.

Log-frequency bins:

\begin{equation}
  f_k = f_{\min} \cdot \left(\frac{f_{\max}}{f_{\min}}\right)^{k/{(K-1)}}
\end{equation}

for $k = 0, 1, \ldots, K-1$ bins.

Alternatively, using equal spacing in log-space:

\begin{equation}
  \log f_k = \log f_{\min} + k \frac{\log f_{\max} - \log f_{\min}}{K - 1}
\end{equation}

\textbf{Linear Interpolation:}

Since FFT bins are linearly spaced, linear interpolation is used
between adjacent bins to get log-spaced values:

\begin{equation}
  y_k = (1 - \alpha) X[i] + \alpha X[i+1]
\end{equation}

where $i = \lfloor f_k / \Delta f \rfloor$ and $\alpha = (f_k / \Delta f) - i$.

\begin{lstlisting}[caption={Log-frequency interpolation via sparse matrix}]

fn build_loghz_matrix(sample_rate_hz: f64, n_fft: usize,
                      n_bins: usize, f_min: f64,
                      f_max: f64) -> SparseMatrix {
    let df = sample_rate_hz / n_fft as f64;

    // Logarithmically-spaced frequencies
    let log_f_min = f_min.ln();
    let log_f_max = f_max.ln();
    let log_step = (log_f_max - log_f_min) / (n_bins - 1) as f64;

    let log_frequencies: Vec<_> = (0..n_bins)
        .map(|i| (i as f64 * log_step + log_f_min).exp())
        .collect();

    // Build interpolation matrix
    let mut matrix = SparseMatrix::new(n_bins, out_len);

    for (bin_idx, &target_freq) in log_frequencies.iter().enumerate() {
        let exact_bin = target_freq / df;
        let lower_bin = exact_bin.floor() as usize;
        let upper_bin = exact_bin.ceil() as usize;

        if lower_bin == upper_bin {
            matrix.set(bin_idx, lower_bin, 1.0);
        } else {
            // Linear interpolation
            let frac = exact_bin - lower_bin as f64;
            matrix.set(bin_idx, lower_bin, 1.0 - frac);
            matrix.set(bin_idx, upper_bin, frac);
        }
    }

    matrix
}
\end{lstlisting}

This matrix is extremely sparse: only 1--2 non-zero values per row
(>99\% sparse).

\subsection{Filterbank Design and Sparse Matrices}

All perceptual scales use sparse matrix multiplication for
efficiency. The operation is:

\begin{equation}
  \mathbf{y} = H \mathbf{x}
\end{equation}

where $H$ is the filterbank matrix (sparse), $\mathbf{x}$ is the
linear spectrum, and $\mathbf{y}$ is the perceptually-scaled spectrum.

\textbf{Sparse Matrix Storage:}

The library uses row-wise sparse format:

\begin{lstlisting}[caption={Row-wise sparse matrix storage definition}]
struct SparseMatrix {
    nrows: usize,
    ncols: usize,
    values: Vec<Vec<f64>>,    // Non-zero values per row
    indices: Vec<Vec<usize>>, // Column indices per row
}
\end{lstlisting}

Multiplication:

\begin{lstlisting}[caption={Sparse matrix-vector multiplication}]
fn multiply_vec(&self, input: &[f64], out: &mut [f64]) {
    for (row_idx, (row_values, row_indices)) in
        self.values.iter().zip(&self.indices).enumerate() {

        let mut acc = 0.0;
        for (&value, &col_idx) in row_values.iter().zip(row_indices) {
            acc += value * input[col_idx];
        }
        out[row_idx] = acc;
    }
}
\end{lstlisting}

\begin{concept}{Sparse Matrix Computational Complexity}
  For matrix-vector multiplication $\mathbf{y} = A\mathbf{x}$ where
  $A$ is $m \times n$:

  \textbf{Dense matrix:}
  \begin{equation}
    \mathcal{O}(mn) \quad \text{(every element accessed)}
  \end{equation}

  \textbf{Sparse matrix with $\text{nnz}$ non-zeros:}
  \begin{equation}
    \mathcal{O}(\text{nnz}) \quad \text{(only non-zeros computed)}
  \end{equation}

  where $\text{nnz}$ is the number of non-zero elements.

  For filterbanks with uniform sparsity $s$ non-zeros per row:
  \begin{equation}
    \text{nnz} = ms
  \end{equation}

  Example: Mel filterbank with $m=80$, $n=257$, $s=20$ non-zeros/row
  has 1,600 operations vs. 20,560 for dense multiplication.
\end{concept}

For mel filterbanks with $M=80$ bands and $N=512$ FFT (257 bins):
\begin{itemize}
  \item Dense matrix: $80 \times 257 = 20{,}560$ multiplications
  \item Sparse matrix: $\approx 80 \times 20 = 1{,}600$
    multiplications ($\thicksim$10--20 non-zeros/row)
\end{itemize}

For log-frequency with linear interpolation:
\begin{itemize}
  \item Only 2 multiplications per output bin (vs. 257 for dense)
\end{itemize}

\section{Constant-Q Transform (CQT)}\label{sec:cqt}

\subsection{Motivation and Q Factor}\label{subsec:cqt_motivation}

The Short-Time Fourier Transform (\cref{subsec:stft_definition})
provides constant frequency resolution $\Delta f$ across all
frequencies. But musical perception is logarithmic: \emph{relative}
frequency differences (ratios) are more relevant than absolute differences.

The \termdef{Constant-Q Transform}
(CQT)~\cite{brown1991calculation,brown1992efficient} provides
logarithmically-spaced frequency bins with constant quality factor $Q$:

\begin{equation}
  \label{eq:q_factor}
  Q = \frac{f_k}{\Delta f_k}
\end{equation}

where $f_k$ is the center frequency and $\Delta f_k$ is the bandwidth
of bin $k$.

For musical applications with $B$ bins per octave:

\begin{equation}
  Q = \frac{1}{2^{1/B} - 1}
\end{equation}

For example, with $B=12$ bins/octave (semitones):
\begin{equation}
  Q = \frac{1}{2^{1/12} - 1} \approx 16.82
\end{equation}

Higher $Q$ → narrower bandwidth → better frequency resolution but
worse time resolution.

\subsection{Kernel-Based Implementation}

Unlike STFT which uses a fixed-length window, CQT uses
\textit{variable-length windows} (kernels) for each frequency:

For frequency bin $k$ with center frequency $f_k$:

\begin{equation}
  N_k = \left\lceil \frac{Q \cdot f_s}{f_k} \right\rceil
\end{equation}

where $f_s$ is the sample rate. Higher frequencies → shorter windows,
lower frequencies → longer windows.

The CQT kernel for frequency $f_k$:

\begin{equation}
  g_k[n] = w[n] \cdot e^{j2\pi f_k n/f_s}, \quad n = 0, 1, \ldots, N_k-1
\end{equation}

where $w[n]$ is a window function (typically Hanning or Hamming).

\textbf{Implementation:}

\begin{lstlisting}[caption={CQT kernel with windowed complex exponential}]

fn generate_kernel_bin(center_freq: f64, kernel_length: usize,
                       sample_rate: f64,
                       window_type: WindowType) -> Vec<Complex<f64>> {
    let mut kernel = Vec::with_capacity(kernel_length);

    // Generate window
    let window = make_window(window_type, kernel_length)?;

    // Generate complex exponential kernel
    for (n, w) in window.iter().enumerate() {
        let t = n as f64 / sample_rate;
        let phase = 2.0 * PI * center_freq * t;

        // e^(i*2*$\pi$*f*t) = cos(2$\pi$ft) + i*sin(2$\pi$ft)
        let exponential = Complex::new(phase.cos(), phase.sin());

        // Apply window function
        let windowed = exponential * w;
        kernel.push(windowed);
    }

    kernel
}
\end{lstlisting}

\subsection{Variable-Length Windows}

The beauty of CQT is that window length scales with frequency:

\begin{itemize}
  \item \textbf{Low frequencies:} Long windows → good frequency
    resolution, poor time resolution
  \item \textbf{High frequencies:} Short windows → good time
    resolution, adequate frequency resolution
\end{itemize}

This matches musical perception: precise pitch resolution is required for low
notes (fundamental frequencies), while less precision is typically acceptable
for high harmonics.

Example for $f_s = 44{,}100$ Hz, $Q \approx 16.82$:
\begin{itemize}
  \item $f = 55$ Hz (A1): $N \approx 13{,}480$ samples ($\approx$305 ms)
  \item $f = 440$ Hz (A4): $N \approx 1{,}685$ samples ($\approx$38 ms)
  \item $f = 3{,}520$ Hz (A7): $N \approx 211$ samples ($\approx$4.8 ms)
\end{itemize}

\subsection{Sparsity Optimization}

CQT kernels naturally have many small-magnitude coefficients,
especially near the edges due to windowing. This can be exploited with
\textit{sparsity thresholding}:

\begin{equation}
  g_k[n] \leftarrow
  \begin{cases}
    g_k[n] & \text{if } |g_k[n]| > \epsilon \cdot \max|g_k| \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}

where $\epsilon$ is the sparsity threshold (e.g., 0.01 for 1\%).

\begin{lstlisting}[caption={Zeroing small CQT kernel coefficients}]

fn apply_sparsity_threshold(kernel: &mut [Complex<f64>],
                            threshold: f64) {
    let max_magnitude = kernel.iter()
                              .map(|c| c.norm())
                              .fold(0.0, f64::max);

    let absolute_threshold = max_magnitude * threshold;

    for coefficient in kernel.iter_mut() {
        if coefficient.norm() < absolute_threshold {
            *coefficient = Complex::new(0.0, 0.0);
        }
    }
}
\end{lstlisting}

Typical sparsity levels: 60--80\% of kernel coefficients can be zeroed
with $\epsilon = 0.01$ with negligible quality loss.

\textbf{Computational benefit:} Sparse kernels reduce the number of
operations in time-domain convolution by skipping zero multiplications.

\begin{concept}{Sparsity Threshold Artifacts}
  \textbf{Warning:} Aggressive sparsity thresholds (e.g., $\epsilon >
  0.01$ or zeroing coefficients below $10^{-10}$) can introduce
  audible artifacts:

  \begin{itemize}[itemsep=2pt]
    \item \textbf{Spectral leakage:} Truncating kernel tails creates
      discontinuities, causing frequency bleeding between adjacent bins.
    \item \textbf{Musical noise:} Sparse, time-varying kernels can
      produce tonal artifacts that vary frame-to-frame, perceived as
      ``chirping'' or ``burbling'' sounds.
    \item \textbf{Time-smearing:} Excessively truncated kernels lose
      their designed frequency selectivity.
  \end{itemize}

  \textbf{Recommended practice:} Use conservative thresholds ($\epsilon
  \leq 0.001$) for high-quality audio analysis. For real-time or
  resource-constrained applications, validate perceptual quality
  through listening tests before deploying aggressive sparsity.
\end{concept}

\section{Advanced Audio Features}\label{sec:advanced}

\subsection{Mel-Frequency Cepstral Coefficients (MFCC)}\label{subsec:mfcc}

MFCCs are the most widely used features in speech recognition,
speaker identification, and audio
classification~\cite{davis1980comparison,rabiner1993fundamentals}.
Introduced by Davis and Mermelstein in 1980, they provide a compact
representation of the spectral envelope.

\textbf{Computation pipeline:}

\begin{enumerate}
  \item Compute mel-scaled power spectrogram: $S_{\text{mel}}[m, k]$
  \item Apply logarithm: $L[m, k] = \log(S_{\text{mel}}[m, k] + \epsilon)$
  \item Apply Discrete Cosine Transform (DCT-II) to each frame
  \item Optionally apply cepstral liftering
  \item Extract first $C$ coefficients (typically 13)
\end{enumerate}

\textbf{Why DCT?} The mel spectrogram has correlated bins due to
overlapping triangular filters. The \termdef{Discrete Cosine
Transform} (DCT)~\cite{ahmed2006discrete} decorrelates them and
compacts energy into the first few coefficients.

\begin{concept}{Discrete Cosine Transform (DCT-II)}
  The DCT-II formula:
  \begin{equation}
    \label{eq:dct_ii}
    c[k] = \sum_{n=0}^{N-1} x[n] \cos\left(\frac{\pi k(n +
    0.5)}{N}\right), \quad k = 0, 1, \ldots, N-1
  \end{equation}

  The DCT is closely related to the DFT but operates entirely on real
  numbers. It's used in JPEG image compression and many audio codecs
  due to its excellent energy compaction properties.
\end{concept}

\begin{lstlisting}[caption={DCT-II implementation for MFCC cepstral step}]

fn dct_ii(input: &[f64]) -> Vec<f64> {
    let n = input.len();
    let mut output = vec![0.0; n];

    for k in 0..n {
        let mut sum = 0.0;
        for (i, &val) in input.iter().enumerate() {
            sum += val * (PI * k as f64 * (i as f64 + 0.5) / n as f64).cos();
        }
        output[k] = sum;
    }

    output
}
\end{lstlisting}

\textbf{Cepstral liftering:}

Liftering applies a sinusoidal weighting to emphasize mid-range coefficients:

\begin{equation}
  \tilde{c}[k] = c[k] \left(1 + \frac{L}{2}\sin\left(\frac{\pi
  k}{L}\right)\right)
\end{equation}

where $L$ is the lifter parameter (commonly 22).

\begin{lstlisting}[caption={Sinusoidal liftering applied to MFCC array}]

fn apply_liftering(mfcc: &mut Array2<f64>, lifter: usize) {
    let n_mfcc = mfcc.nrows();

    // Compute lifter weights
    let weights: Vec<_> = (0..n_mfcc)
        .map(|i| 1.0 + (lifter as f64 / 2.0)
                     * (PI * i as f64 / lifter as f64).sin())
        .collect();

    // Apply weights to each frame
    for frame_idx in 0..n_frames {
        for coeff_idx in 0..n_mfcc {
            mfcc[[coeff_idx, frame_idx]] *= weights[coeff_idx];
        }
    }
}
\end{lstlisting}

\textbf{The C0 coefficient:} The zeroth coefficient represents the
overall energy of the frame. Some applications include it (speaker
recognition), others exclude it (speech recognition) to reduce
sensitivity to volume.

\textbf{Why 13 coefficients?} Empirically, 13 MFCCs capture most
speech information. Higher coefficients represent fine spectral
detail that's often noise.

\subsection{Chroma Features for Music Analysis}\label{subsec:chroma}

Chroma features~\cite{fujishima1999realtime,muller2005audio} (also
called pitch class profiles) represent the energy distribution across
the 12 pitch classes of the Western musical scale, collapsing all
octaves together.

The 12 pitch classes: C, C\musSharp{}, D, D\musSharp{}, E, F,
F\musSharp{}, G, G\musSharp{}, A, A\musSharp{}, B

\textbf{Construction:}

\begin{enumerate}
  \item Map FFT bins to musical notes using:
    \begin{equation}
      n(f) = 12 \log_2\left(\frac{f}{f_{\text{ref}}}\right)
    \end{equation}
    where $f_{\text{ref}}$ is a reference frequency (e.g., A4 = 440 Hz)

  \item Determine pitch class: $p = n \mod 12$

  \item Sum energy from all octaves for each pitch class:
    \begin{equation}
      C[p] = \sum_{\text{octaves}} E[p, \text{octave}]
    \end{equation}

  \item Normalize (L1, L2, or max)
\end{enumerate}

\textbf{Normalization strategies:}

\begin{itemize}
  \item \textbf{L1:} $\tilde{C}[p] = C[p] / \sum_{p'} C[p']$ (probabilities)
  \item \textbf{L2:} $\tilde{C}[p] = C[p] / \sqrt{\sum_{p'} C[p']^{2}}$
    (Euclidean norm)
  \item \textbf{Max:} $\tilde{C}[p] = C[p] / \max_{p'} C[p']$
    (relative strength)
\end{itemize}

\textbf{Applications:} Chord recognition, key estimation, cover song
identification, music similarity.

\subsection{ERB Spectrograms and Gammatone Filters}

ERB spectrograms using gammatone filters provide the most
perceptually accurate time-frequency representation.

The gammatone impulse response:

\begin{equation}
  g(t) = at^{n-1}e^{-2\pi bt}\cos(2\pi f_c t + \phi)
\end{equation}

where:
\begin{itemize}
  \item $a$ is amplitude
  \item $n$ is filter order (typically 4)
  \item $b$ is bandwidth parameter
  \item $f_c$ is center frequency
  \item $\phi$ is phase
\end{itemize}

The bandwidth parameter relates to ERB:\

\begin{equation}
  b = 1.019 \cdot \text{ERB}(f_c)
\end{equation}

\textbf{Frequency domain implementation:}

For efficiency, the library implements gammatone filtering in the
frequency domain. The frequency-domain transfer function is:

\begin{equation}
  G(f) = \frac{1}{\left(1 + j\frac{f - f_c}{b}\right)^n}
\end{equation}

where $n$ is the filter order (typically 4), $f_c$ is the center
frequency, $b$ is the bandwidth parameter, and $j = \sqrt{-1}$.
This fourth-order complex pole provides the characteristic asymmetric
frequency response of the auditory filter.

Filtering is then performed via element-wise multiplication:

\begin{equation}
  Y(f) = X(f) \cdot G(f)
\end{equation}

This avoids expensive time-domain convolution for each filter.

\section{Image Processing via FFT}\label{sec:image}

\subsection{2D Convolution and the Convolution
Theorem}\label{subsec:convolution_theorem}

The \termdef{convolution theorem}~\cite{blozinski1972convolution} states
that convolution in the spatial domain is equivalent to
multiplication in the frequency domain:

\begin{equation}
  \label{eq:convolution_theorem}
  f \ast g = \mathcal{F}^{-1}\{\mathcal{F}\{f\} \cdot \mathcal{F}\{g\}\}
\end{equation}

For images:
\begin{equation}
  (I \ast K)[x, y] = \text{IFFT2D}\{\text{FFT2D}(I) \odot \text{FFT2D}(K)\}
\end{equation}

where $\odot$ denotes element-wise multiplication.

\textbf{Complexity comparison:}

Spatial convolution for image $M \times N$ with kernel $K \times K$:
\begin{equation}
  \mathcal{O}(M \times N \times K^2)
\end{equation}

FFT-based convolution:
\begin{equation}
  \mathcal{O}(M N \log(MN))
\end{equation}

\textbf{Theoretical crossover:} FFT has better asymptotic complexity when $K >
\sqrt{\log(MN)}$, typically around $K \geq 7$.

\subsection{Kernel Padding and Phase Shifting}

For correct FFT convolution, the kernel must be padded to image size
with proper phase shifting. The kernel's center should map to
position $(0, 0)$ in the padded array:

\begin{lstlisting}[caption={Padding kernel with centre at origin for FFT}]

fn pad_kernel_for_fft(kernel: &Array2<f64>,
                      target_shape: (usize, usize)) -> Array2<f64> {
    let (ker_rows, ker_cols) = kernel.dim();
    let mut result = Array2::<f64>::zeros(target_shape);

    // Kernel center position
    let ker_center_row = ker_rows / 2;
    let ker_center_col = ker_cols / 2;

    // Place kernel with center at (0, 0) using wraparound
    for i in 0..ker_rows {
        for j in 0..ker_cols {
            let row_offset = i as isize - ker_center_row as isize;
            let col_offset = j as isize - ker_center_col as isize;

            // Wrap around to place center at (0, 0)
            let target_row = row_offset.rem_euclid(target_rows as isize) as usize;
            let target_col = col_offset.rem_euclid(target_cols as isize) as usize;

            result[[target_row, target_col]] = kernel[[i, j]];
        }
    }

    result
}
\end{lstlisting}

This ensures the FFT interprets the kernel correctly for circular convolution.

\subsection{Spatial Filtering in the Frequency Domain}

Filtering is multiplication by a frequency-domain mask:

\begin{equation}
  Y(u, v) = X(u, v) \cdot H(u, v)
\end{equation}

\textbf{Low-pass filter:} Keeps low frequencies, removes high
frequencies (smoothing)

\begin{equation}
  H_{\text{LP}}(u, v) =
  \begin{cases}
    1 & \sqrt{u^2 + v^2} \leq D_0 \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}

\textbf{High-pass filter:} Removes low frequencies, keeps high
frequencies (sharpening/edges)

\begin{equation}
  H_{\text{HP}}(u, v) = 1 - H_{\text{LP}}(u, v)
\end{equation}

\textbf{Band-pass filter:} Keeps frequencies in a range

\begin{equation}
  H_{\text{BP}}(u, v) =
  \begin{cases}
    1 & D_1 \leq \sqrt{u^2 + v^2} \leq D_2 \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}

\subsection{Gaussian Blur via FFT}

The Gaussian kernel for blurring:

\begin{equation}
  G(x, y) = \frac{1}{2\pi\sigma^2} \exp\left(-\frac{x^2 + y^2}{2\sigma^2}\right)
\end{equation}

Implementation:

\begin{lstlisting}[caption={2D Gaussian kernel generation and normalisation}]

pub fn gaussian_kernel_2d(size: NonZeroUsize, sigma: f64) -> SpectrogramResult<Array2<f64>> {
    if size.get().is_multiple_of(2) {
        return Err(SpectrogramError::invalid_input(
            "kernel size must be odd and > 0",
        ));
    }
    if sigma <= 0.0 {
        return Err(SpectrogramError::invalid_input("sigma must be > 0"));
    }
    let size = size.get();
    let center = (size / 2) as f64;
    let variance = sigma * sigma;
    let coeff = 1.0 / (2.0 * PI * variance);

    let mut kernel = Array2::<f64>::zeros((size, size));

    for i in 0..size {
        for j in 0..size {
            let x = i as f64 - center;
            let y = j as f64 - center;
            let exponent = -(x * x + y * y) / (2.0 * variance);
            kernel[[i, j]] = coeff * exponent.exp();
        }
    }

    // Normalize to sum to 1.0
    let sum: f64 = kernel.iter().sum();
    kernel.mapv_inplace(|v| v / sum);

    Ok(kernel)
}
\end{lstlisting}

The Gaussian has a useful property: its Fourier transform is also
Gaussian, ensuring smooth frequency response.

\subsection{Edge Detection and High-Pass Filtering}

Edges correspond to high-frequency components. A simple high-pass
filter enhances edges:

\begin{equation}
  \text{Edges} = \text{Image} - \text{Lowpass}(\text{Image})
\end{equation}

Or apply a high-pass mask directly in frequency domain:

\begin{equation}
  H_{\text{HP}}(u, v) =
  \begin{cases}
    0 & \sqrt{u^2 + v^2} < D_0 \\
    1 & \text{otherwise}
  \end{cases}
\end{equation}

The Laplacian operator for edge detection:

\begin{equation}
  \nabla^2 I = \frac{\partial^2 I}{\partial x^2} + \frac{\partial^2
  I}{\partial y^2}
\end{equation}

In frequency domain, differentiation becomes multiplication:

\begin{equation}
  \mathcal{F}\{\nabla^2 I\} = -(u^2 + v^2) \mathcal{F}\{I\}
\end{equation}

\section{Binaural Spectrograms}\label{sec:binaural}

\subsection{Spatial Hearing and Binaural Cues}\label{subsec:binaural_intro}

The human auditory system localises sound by comparing the signals
arriving at the two ears.\ The library exposes four classic binaural
cues as time-frequency representations, each implemented as a
spectrogram whose rows correspond to frequency bins and whose columns
correspond to STFT frames.\ All four operate at either \code{f32} or
\code{f64} precision.

\begin{center}
  \begin{tabular}{llll}
    \textbf{Cue} & \textbf{Symbol} & \textbf{Units} &
    \textbf{Frequency range} \\
    \hline
    Interaural Time Difference & ITD & seconds & $<$1.5~kHz \\
    Interaural Phase Difference & IPD & radians & all \\
    Interaural Level Difference & ILD & dB & $>$1.5~kHz \\
    Interaural Level Ratio & ILR & dimensionless ($[-1,1]$) & all \\
  \end{tabular}
\end{center}

\subsection{Interaural Time Difference (ITD)}\label{subsec:itd}

\begin{concept}{ITD Definition}
  The ITD is the extra propagation time for a sound to reach the
  farther ear:

  \begin{equation}
    \label{eq:itd}
    \text{ITD}(k, m) = \frac{\angle X_L[k,m] - \angle X_R[k,m]}{2\pi f_k}
  \end{equation}

  where $X_L$ and $X_R$ are the complex STFT spectra of the left and
  right channels, $\angle$ denotes argument (phase angle), and $f_k$
  is the centre frequency of bin $k$ in Hz.\ At low frequencies the
  phase difference is an unambiguous indicator of ITD; above $\sim$1.5~kHz
  phase wrapping makes the relationship ambiguous.
\end{concept}

The library uses \code{round} (not \code{floor}/\code{ceil}) for the
bin-index conversion to match the \texttt{Binaspect} reference
implementation, and returns only the bins spanning
\code{[start\_freq,~end\_freq]} rather than the full spectrum:

\begin{lstlisting}[caption={Computing ITD spectrogram from stereo channels}]

use spectrograms::{ITDSpectrogramParams, compute_itd_spectrogram};
use std::num::NonZeroUsize;

let n_fft = NonZeroUsize::new(4096).unwrap();
let hop   = NonZeroUsize::new(2048).unwrap();
let params = ITDSpectrogramParams::new(50.0, 620.0, n_fft, hop)?;
// start_freq=50 Hz, end_freq=620 Hz

let itd = compute_itd_spectrogram(&left, &right, sample_rate, &params)?;
// itd is an ItdSpectrogram<f64>; shape is (n_freq_bins, n_frames)
\end{lstlisting}

\subsection{Interaural Phase Difference (IPD)}\label{subsec:ipd}

The IPD is the unwrapped phase difference between channels:

\begin{equation}
  \text{IPD}(k, m) = \angle X_L[k,m] - \angle X_R[k,m]
\end{equation}

Values lie in $(-\pi, \pi]$. Unlike ITD, IPD is frequency-independent
and does not require dividing by $f_k$, making it ambiguous at high
frequencies (half a wavelength becomes less than the inter-ear distance).

\subsection{Interaural Level Difference (ILD)}\label{subsec:ild}

\begin{concept}{ILD Definition}
  The ILD captures level asymmetry between channels in decibels:

  \begin{equation}
    \label{eq:ild}
    \text{ILD}(k, m) = 10 \log_{10}
    \frac{|X_L[k,m]|^2 + \epsilon}{|X_R[k,m]|^2 + \epsilon}
  \end{equation}

  where $\epsilon$ prevents division by zero.\ Positive values indicate
  the sound source is closer to the left ear.\ ILD is the dominant
  cue above $\sim$1.5~kHz because the head creates a significant
  acoustic shadow at those wavelengths.
\end{concept}

\subsection{Interaural Level Ratio (ILR)}\label{subsec:ilr}

The ILR normalises the power difference to the $[-1,1]$ range:

\begin{equation}
  \label{eq:ilr}
  \text{ILR}(k, m) = \frac{|X_L[k,m]|^2 - |X_R[k,m]|^2}
  {|X_L[k,m]|^2 + |X_R[k,m]|^2 + \epsilon}
\end{equation}

A value of $+1$ indicates all energy in the left channel; $-1$
indicates all energy in the right channel; $0$ is perfectly balanced.

\subsection{Histogram Analysis and Comparison}\label{subsec:binaural_histogram}

Every binaural spectrogram object exposes a \code{histogram(n\_bins)}
method that aggregates the frame-by-frame cue values into a 2D
histogram (frequency bins $\times$ value buckets), providing a
statistical summary of the spatial distribution over time.

The library also provides comparison functions that compute the
mean per-frequency-bin difference between a reference and a test
recording:

\begin{lstlisting}[caption={Mean ITD difference between reference and test}]

let (col_means, mean_diff_degrees, mean_diff_itd) =
    compute_itd_spectrogram_diff(&ref_left, &ref_right,
                                 &test_left, &test_right,
                                 sample_rate, &params)?;
// col_means         -- per-frequency-bin mean ITD difference (Array1)
// mean_diff_degrees -- mean absolute error converted to degrees of azimuth
// mean_diff_itd     -- median ITD difference across frames (seconds)
\end{lstlisting}

\subsection{Parallelisation}\label{subsec:binaural_rayon}

When the \code{rayon} feature is enabled, the per-frequency-bin
magphase computation uses \code{par\_for\_each}, giving a significant
speed-up on multi-core systems.\ Enable it in \texttt{Cargo.toml}:

\begin{lstlisting}[caption={Enabling rayon parallelisation in Cargo.toml}]

[dependencies]
spectrograms = { version = "2", features = ["rayon"] }
\end{lstlisting}

\section{Modified Discrete Cosine Transform (MDCT)}\label{sec:mdct}

\subsection{Motivation and Codec Context}\label{subsec:mdct_motivation}

The \termdef{Modified Discrete Cosine Transform} (MDCT) is a lapped
orthogonal transform at the heart of virtually every modern lossy audio
codec: MP3 (Layer III), AAC, Vorbis, and Opus all use it as their
core time-frequency analysis stage.\ Unlike the STFT, the MDCT
overlaps input frames by exactly 50\% and maps $2N$ samples to $N$
real-valued coefficients (50\% dimensionality reduction), yet still
achieves \termdef{perfect reconstruction} when the appropriate window
is applied.

\subsection{MDCT Definition}\label{subsec:mdct_definition}

For a windowed frame of $2N$ samples $x[n]$, the MDCT produces $N$
coefficients:

\begin{equation}
  \label{eq:mdct}
  C[k] = \sum_{n=0}^{2N-1} x[n]\, w[n]\, \cos\!\left(
    \frac{\pi\,(2n+1+N)\,(2k+1)}{4N}
  \right), \quad k = 0, 1, \ldots, N-1
\end{equation}

where $w[n]$ is a window function satisfying the TDAC condition
(see \cref{subsec:mdct_pr}).

The inverse MDCT (IMDCT) reconstructs $2N$ samples from $N$ coefficients:

\begin{equation}
  \label{eq:imdct}
  y[n] = \frac{2}{N} \sum_{k=0}^{N-1} C[k]\, w[n]\, \cos\!\left(
    \frac{\pi\,(2n+1+N)\,(2k+1)}{4N}
  \right)
\end{equation}

Consecutive frames overlap by $N$ samples and are summed
(\termdef{overlap-add}); the TDAC window ensures cancellation of
aliasing terms.

\subsection{Fast Computation via the Packing Trick}\label{subsec:mdct_fft}

Direct evaluation of \cref{eq:mdct} costs $O(N^2)$.\ The library
computes the forward MDCT in $O(N \log N)$ using a single $N$-point
complex-to-complex (C2c) FFT (the \emph{packing trick}):

\begin{enumerate}
  \item Pre-multiply each windowed sample by a twiddle factor to produce
    a complex sequence $z[m] = a[m] + jb[m]$ of length $N$ from the
    $2N$-sample frame.
  \item Execute one $N$-point C2c FFT on $z$.
  \item Because $a$ and $b$ are real, the result satisfies Hermitian
    symmetry; the $2N$-point DFT at even and odd indices is recovered via:
    \begin{equation}
      F[2p] = Z[p], \qquad F[2p+1] = Z[N-1-p]^*
    \end{equation}
  \item Apply a post-twiddle to obtain the MDCT coefficients.
\end{enumerate}

This halves the FFT count compared to the naive 2$\times$R2c($N$) approach.

The inverse (IMDCT) similarly uses 2$\times$C2c($N$) FFTs by splitting
the half-sparse $2N$-point DFT into even- and odd-indexed subsequences.

\subsection{Perfect Reconstruction and the TDAC Condition}\label{subsec:mdct_pr}

\begin{concept}{Time-Domain Aliasing Cancellation (TDAC)}
  Perfect reconstruction requires the window $w$ to satisfy:
  \begin{equation}
    \label{eq:tdac}
    w[n]^2 + w[n+N]^2 = 1, \quad \forall\, n \in [0, N)
  \end{equation}

  \textbf{Only the sine window satisfies this condition} for 50\% hop:
  \begin{equation}
    \label{eq:sine_window}
    w[n] = \sin\!\left(\frac{\pi (n + \tfrac{1}{2})}{2N}\right)
  \end{equation}

  Standard windows (Hanning, Hamming, Blackman, Kaiser) do \emph{not}
  satisfy \cref{eq:tdac} and will introduce reconstruction error.
\end{concept}

\textbf{Usage:}

\begin{lstlisting}[caption={MDCT and IMDCT with sine-window reconstruction}]

use spectrograms::{MdctParams, mdct, imdct};
use std::num::NonZeroUsize;

// Sine window gives perfect reconstruction with 50% hop.
let params = MdctParams::sine_window(NonZeroUsize::new(2048).unwrap())?;

let frames: Vec<Vec<f64>> = /* overlapping 2048-sample frames */;
let coeffs = mdct(&frames, &params)?;     // (n_frames, 1024) coefficients

let reconstructed = imdct(&coeffs, &params)?;  // overlap-added output
\end{lstlisting}

\textbf{Single precision:}

\begin{lstlisting}[caption={MDCT at single-precision f32 via turbofish}]

let coeffs = mdct::<f32>(&frames_f32, &params)?;
\end{lstlisting}

\subsection{Parameters}\label{subsec:mdct_params}

\begin{center}
  \begin{tabular}{lll}
    \textbf{Parameter} & \textbf{Type} & \textbf{Notes} \\
    \hline
    \code{window\_size} & \code{NonZeroUsize} (even, $\geq 4$) &
    Total frame length $2N$ \\
    \code{hop\_size} & \code{NonZeroUsize} & Typically $N$ for PR \\
    \code{window} & \code{WindowType} &
    Use \code{MdctParams::sine\_window} for PR \\
  \end{tabular}
\end{center}

\section{FFT Convolution and Minimum Phase}\label{sec:convolution}

\subsection{FFT-Based Linear Convolution}\label{subsec:fft_convolve}

The convolution theorem (\cref{subsec:convolution_theorem}) applies
equally to 1D real signals.\ For two signals $a$ of length $M$ and $b$
of length $K$, linear convolution yields an output of length $M + K - 1$.
Direct time-domain convolution costs $O(MK)$; the FFT-based approach
costs $O((M+K) \log(M+K))$ and is more efficient whenever $K \gtrsim
\sqrt{\log(M+K)}$.

\begin{lstlisting}[caption={FFT linear convolution with NonEmptySlice input}]

use spectrograms::fft_convolve;
use non_empty_slice::NonEmptySlice;

let signal: Vec<f64> = /* ... */;
let kernel: Vec<f64> = /* FIR filter taps */;

let sig_ne = NonEmptySlice::new(&signal).unwrap();
let ker_ne = NonEmptySlice::new(&kernel).unwrap();

let output = fft_convolve(sig_ne, ker_ne)?;
// output.len() == signal.len() + kernel.len() - 1
\end{lstlisting}

\textbf{Algorithm:} Both inputs are zero-padded to the next
power-of-two above $M + K - 1$, multiplied element-wise in the
frequency domain (a single complex multiplication per frequency bin),
and inverse-transformed.

\subsection{Spectral-Division Deconvolution}\label{subsec:fft_deconvolve}

Given an observed output $y = h \ast x$ and the filter $h$, the
original signal $x$ can be recovered by dividing spectra.\ Exact
spectral division is unstable at frequencies where $H(f) \approx 0$;
\termdef{regularised deconvolution} addresses this:

\begin{equation}
  \hat{X}(f) = \frac{Y(f)\, \conj{H}(f)}{|H(f)|^2 + \varepsilon}
\end{equation}

where $\varepsilon = \lambda \cdot \max_f |H(f)|^2$ and $\lambda$ is
the \code{regularization} parameter.\ With $\lambda = 0$ this reduces
to exact division; small positive values stabilise the result near
spectral nulls.

\begin{lstlisting}[caption={Regularised spectral deconvolution}]

use spectrograms::fft_deconvolve;

let recovered = fft_deconvolve(output_ne, kernel_ne, 1e-6)?;
\end{lstlisting}

\subsection{Streaming Overlap-Save Convolution}\label{subsec:overlap_save}

For real-time or block-streaming applications (e.g., room-correction
FIR filtering in an audio callback), repeatedly calling
\code{fft\_convolve} is wasteful because the impulse response $h$
is fixed and its FFT could be cached.\ The
\code{OverlapSaveConvolver} does exactly this.

\begin{concept}{Overlap-Save Method}
  At construction, the impulse response $h$ of length $K$ is
  zero-padded to the FFT size $N_{\text{FFT}} =
  \text{next\_power\_of\_two}(B + K - 1)$ and transformed once.
  For each input block of length $B$:
  \begin{enumerate}
    \item Prepend the $N_{\text{FFT}} - B$ most recent samples
      (\emph{history}) to the block to form a window of length $N_{\text{FFT}}$.
    \item Compute an $N_{\text{FFT}}$-point C2c FFT.
    \item Multiply element-wise by the cached $H$ spectrum.
    \item Inverse-transform.
    \item Discard the first $K - 1$ alias-contaminated samples; the
      remaining $B$ samples are the alias-free output.
  \end{enumerate}

  Per-block cost: $O(B \log N_{\text{FFT}})$, compared to $O(BK)$ for
  direct convolution.\ \textbf{No heap allocation on the processing path.}
\end{concept}

\begin{lstlisting}[caption={Streaming overlap-save with fixed block size}]

use spectrograms::OverlapSaveConvolver;
use std::num::NonZeroUsize;

// Build once from the impulse response.
let conv = OverlapSaveConvolver::new(&ir, NonZeroUsize::new(256).unwrap())?;

// In the audio callback (no allocation):
let mut output = vec![0.0f64; 256];
conv.process_block(&input_block, &mut output)?;
\end{lstlisting}

Key properties:
\begin{itemize}
  \item Fixed block size set at construction; every call to
    \code{process\_block} must provide exactly that many samples.
  \item Call \code{reset()} between independent streams to clear the
    inter-block history.
  \item Generic over \code{T: Sample} (works at both \code{f32} and
    \code{f64}).
\end{itemize}

\subsection{Minimum-Phase Conversion}\label{subsec:minimum_phase}

A \termdef{minimum-phase} filter has the same magnitude response as a
given (typically linear-phase) FIR filter, but concentrates its energy
at the start rather than the middle.\ This eliminates most of the
filter's latency---beneficial for live-processing and room-correction
applications.

\begin{concept}{Real-Cepstrum Method}
  The library uses the homomorphic (real-cepstrum) algorithm:
  \begin{enumerate}
    \item Compute $H = \text{FFT}(h)$ on a heavily oversampled grid
      (default: $8\times$ the IR length, rounded up to a power of two)
      to avoid cepstral time-aliasing.
    \item Form the real log-magnitude: $\hat{H} = \log|H|$.
    \item Apply an inverse FFT to obtain the real cepstrum $c$.
    \item Window $c$: keep the DC and Nyquist taps unchanged,
      double all causal taps (indices 1 to $N/2 - 1$), and zero all
      anticausal taps (indices $N/2 + 1$ to $N - 1$).
      This is the discrete version of the Hilbert relation that derives
      minimum-phase angle from log-magnitude.
    \item Compute $H_{\min} = \exp(\text{FFT}(c))$, then
      $h_{\min} = \text{Re}(\text{IFFT}(H_{\min}))$.
    \item Truncate to the desired output length.
  \end{enumerate}

  The magnitude response of $h_{\min}$ matches that of $h$ to within
  the accuracy of the oversampled FFT; increasing the oversample factor
  reduces the residual magnitude error at the cost of a larger one-off
  transform.
\end{concept}

\begin{lstlisting}[caption={Converting linear-phase FIR to minimum phase}]

use spectrograms::minimum_phase;

// A 64-tap linear-phase low-pass prototype.
let lin_phase_fir: Vec<f32> = /* ... */;
let mp_fir = minimum_phase(&lin_phase_fir)?;

// mp_fir has the same length and (nearly) the same magnitude response,
// but most energy is in the first few taps.
assert!(mp_fir[0].abs() >= mp_fir[mp_fir.len() - 1].abs());
\end{lstlisting}

\textbf{When to use minimum phase.} A linear-phase FIR of length $L$
introduces a group delay of $(L-1)/2$ samples.\ For a 1000-tap room
correction filter at 48~kHz this is over 10~ms of latency.\ The
minimum-phase equivalent reduces this to near zero while preserving the
equalisation curve.\ Use \code{minimum\_phase\_with} to control the
output length (useful when only the leading portion of the minimum-phase
response is needed) and the oversampling factor (higher values give a
more accurate magnitude match).

\section{Computational Optimizations}\label{sec:optimizations}

\subsection{Plan-Based Computation}\label{subsec:fft_plans}

FFT planning involves:
\begin{enumerate}
  \item Analyzing the transform size
  \item Selecting optimal algorithm (radix-2, split-radix, Bluestein, etc.)
  \item Precomputing twiddle factors: $W_N^k = e^{-j2\pi k/N}$
  \item Allocating scratch buffers
\end{enumerate}

For a single FFT, planning overhead is negligible. But for
spectrograms with hundreds of frames, planning each FFT is wasteful.

\textbf{Solution: Reusable plans}

\begin{lstlisting}[caption={Reusable STFT plan with cached FFT and window}]

pub struct StftPlan {
    n_fft: usize,
    hop: usize,
    window: Vec<f64>,      // Pre-computed window
    fft: Box<dyn R2cPlan>, // Reusable FFT plan
    fft_out: Vec<Complex<f64>>, // Reusable buffer
    frame: Vec<f64>,       // Reusable frame buffer
}

impl StftPlan {
    pub fn new(params: &SpectrogramParams) -> Result<Self> {
        let n_fft = params.stft().n_fft();
        let window = make_window(params.stft().window(), n_fft)?;

        // Create FFT plan once
        let mut planner = RealFftPlanner::new();
        let fft = planner.plan_r2c(n_fft)?;

        Ok(Self {
            n_fft,
            window,
            fft,
            fft_out: vec![Complex::new(0.0, 0.0); n_fft/2 + 1],
            frame: vec![0.0; n_fft],
        })
    }

    // Reuse plan for each frame
    fn compute_frame(&mut self, samples: &[f64],
                     frame_idx: usize) -> Result<()> {
        // Window frame into self.frame buffer
        // ...

        // Reuse FFT plan and buffers
        self.fft.process(&self.frame, &mut self.fft_out)?;
        Ok(())
    }
}
\end{lstlisting}

\textbf{Computational benefit:} For 1000-frame spectrograms, plan reuse
amortizes the planning overhead across all frames.

\subsection{Sparse Matrix Operations}

For filterbank multiplication $\mathbf{y} = H\mathbf{x}$ where $H$ is sparse:

Dense: $\mathcal{O}(mn)$ for $m \times n$ matrix

Sparse: $\mathcal{O}(nnz)$ where $nnz$ is number of non-zeros

For mel filterbanks with 80 bands and 257 FFT bins:
\begin{itemize}
  \item Dense: $80 \times 257 = 20{,}560$ ops
  \item Sparse ($\thicksim$20 non-zeros/row): $80 \times 20 = 1{,}600$ ops
\end{itemize}

\textbf{Row-wise sparse storage:}

\begin{lstlisting}[caption={Sparse matrix layout and filterbank multiply}]
struct SparseMatrix {
    nrows: usize,
    ncols: usize,
    values: Vec<Vec<f64>>,     // Non-zero values per row
    indices: Vec<Vec<usize>>,  // Column indices per row
}

fn multiply_vec(&self, input: &[f64], out: &mut [f64]) {
    for (row_idx, (row_values, row_indices)) in
        self.values.iter().zip(&self.indices).enumerate() {

        let mut acc = 0.0;
        for (&value, &col_idx) in
            row_values.iter().zip(row_indices) {
            acc += value * input[col_idx];
        }
        out[row_idx] = acc;
    }
}
\end{lstlisting}

\textbf{Cache efficiency:} Row-wise format ensures sequential access
to both input array and sparse values, maximizing cache hits.

\subsection{ Design Patterns}

Minimize allocations in hot loops:

\textbf{Bad: Allocate per iteration}
\begin{lstlisting}[caption={Per-frame allocation — wasteful hot-loop pattern}]
for frame in 0..n_frames {
    let mut windowed = vec![0.0; n_fft]; // Allocation!
    // ... process frame
}
\end{lstlisting}

\textbf{Good: Reuse workspace}
\begin{lstlisting}[caption={Pre-allocated workspace reused across frames}]
let mut workspace = vec![0.0; n_fft]; // Single allocation
for frame in 0..n_frames {
    // Reuse workspace
    fill_frame(&mut workspace, samples, frame);
    // ... process frame
}
\end{lstlisting}

\textbf{Workspace pattern:}

\begin{lstlisting}[caption={Full workspace struct with resizable buffers}]
struct Workspace {
    frame: Vec<f64>,          // Windowed frame buffer
    fft_out: Vec<Complex<f64>>, // FFT output buffer
    spectrum: Vec<f64>,       // Power spectrum buffer
    mapped: Vec<f64>,         // Filterbank output buffer
}

impl Workspace {
    fn ensure_sizes(&mut self, n_fft: usize,
                   out_len: usize, n_bins: usize) {
        self.frame.resize(n_fft, 0.0);
        self.fft_out.resize(out_len, Complex::new(0.0, 0.0));
        self.spectrum.resize(out_len, 0.0);
        self.mapped.resize(n_bins, 0.0);
    }
}
\end{lstlisting}

\subsection{Memory Layout and Cache Efficiency}

\textbf{Row-major vs.\ column-major:}

The library uses row-major layout (C-style) for compatibility with
most FFT libraries:

\begin{lstlisting}[caption={Asserting row-major layout for FFT compatibility}]
// Row-major: rows are contiguous
let data = Array2::<f64>::zeros((nrows, ncols));
assert!(data.is_standard_layout()); // row-major
\end{lstlisting}

For column-wise operations (processing spectrogram frames), this can
cause cache misses. But the benefit of contiguous memory for FFT outweighs this.

\textbf{Alignment:} Modern FFT libraries (FFTW, RealFFT) benefit from
aligned memory. The library ensures contiguous slices are passed to
FFT routines.

\subsection{FFTW Integration}\label{subsec:fftw}

FFTW (\emph{Fastest Fourier Transform in the
West})~\cite{frigo2005design} is considered the gold standard for FFT
performance. The library supports both:

\begin{itemize}
  \item \textbf{RealFFT:} Pure Rust, portable, good performance
  \item \textbf{FFTW:} C library, excellent performance, requires
    system installation
\end{itemize}

\textbf{FFTW wisdom:}

FFTW learns optimal algorithms through runtime measurements. Plans
can be saved and reused:

\begin{lstlisting}[caption={FFTW planning modes from ESTIMATE to PATIENT}]
// FFTW planning flags:
// ESTIMATE: Fast planning, ok performance
// MEASURE: Slower planning, better performance
// PATIENT: Very slow planning, best performance
let plan = fftw::plan_r2c_1d(n_fft, FFTW_MEASURE);
\end{lstlisting}

For repeated transforms of the same size, MEASURE or PATIENT modes
can provide better performance than ESTIMATE.

\section{Numerical Considerations}\label{sec:numerical}

\subsection{Floating-Point Precision}\label{subsec:precision-f64}

The library defaults to \code{f64} (double precision), and double
precision is the recommended choice for offline analysis:

\begin{itemize}
  \item 53 bits of precision ($\approx$16 decimal digits)
  \item Dynamic range: $\approx 10^{-308}$ to $10^{308}$
  \item Ample headroom for all audio processing needs
\end{itemize}

As described in \cref{subsec:precision}, the core algorithms are
generic over the scalar type and may also be instantiated at \code{f32}.

\textbf{When to prefer f64.} Double precision is advantageous when
audio processing requires:
\begin{itemize}
  \item Accurate accumulation over many samples (STFT frames)
  \item Stable recursive filters (if implemented)
  \item Wide dynamic range for decibel conversions
\end{itemize}

\textbf{When to prefer f32.} Single precision halves memory traffic for
memory-bound workloads, matches machine-learning pipelines that train
in single precision, and suits embedded or \code{no\_std} targets with
single-precision floating-point units. \code{f64} remains the default
because it ensures numerical stability without performance concerns on
modern CPUs.

\subsection{Normalization Conventions}\label{subsec:normalization}

Different conventions exist for FFT normalization. The library uses:

\textbf{Forward FFT:} No normalization
\begin{equation}
  X[k] = \sum_{n=0}^{N-1} x[n] e^{-j2\pi kn/N}
\end{equation}

\textbf{Inverse FFT:} Divide by $N$
\begin{equation}
  x[n] = \frac{1}{N}\sum_{k=0}^{N-1} X[k] e^{j2\pi kn/N}
\end{equation}

This ensures $\text{IFFT}(\text{FFT}(x)) = x$ exactly.

Alternative conventions (used by some libraries):
\begin{itemize}
  \item Both directions scaled by $1/\sqrt{N}$ (symmetric)
  \item Inverse scaled by $1/N$, forward by $N$ (rare)
\end{itemize}

\subsection{Dynamic Range and Epsilon Thresholds}

\textbf{Logarithm floor:}

To avoid $\log(0) = -\infty$:

\begin{lstlisting}[caption={Log floor with epsilon to avoid negative infinity}]
let epsilon = 1e-10;
let db = 10.0 * (power.max(epsilon)).log10();
\end{lstlisting}

Choosing $\epsilon$:
\begin{itemize}
  \item Too large: Loss of quiet signals
  \item Too small: Numerical instability
  \item Sweet spot: $10^{-10}$ to $10^{-8}$ for f64
\end{itemize}

\textbf{Sparse matrix threshold:}

For considering values ``zero'':

\begin{lstlisting}[caption={Epsilon threshold for sparse matrix storage}]
if value.abs() > 1e-10 {
    store(value);
}
\end{lstlisting}

This is much larger than machine epsilon ($\approx 2.2 \times
10^{-16}$ for f64) to account for accumulated rounding errors.

\section{Conclusion and Further Reading}

\subsection{Summary of Key Concepts}

This manual covered:

\textbf{Core algorithms:}
\begin{itemize}
  \item DFT and FFT (Cooley-Tukey algorithm)
  \item Real-to-complex FFT with Hermitian symmetry
  \item 2D FFT via row-column decomposition
  \item STFT and time-frequency uncertainty
\end{itemize}

\textbf{Signal processing methods:}
\begin{itemize}
  \item Window functions (Hanning, Hamming, Blackman, Kaiser, Gaussian)
  \item Amplitude scaling (power, magnitude, decibels)
  \item Convolution theorem for efficient filtering
\end{itemize}

\textbf{Perceptual scales:}
\begin{itemize}
  \item Mel scale for speech processing
  \item ERB scale (Glasberg \& Moore) for auditory modeling
  \item Logarithmic frequency for music
  \item CQT with variable-length kernels
\end{itemize}

\textbf{Advanced features:}
\begin{itemize}
  \item MFCCs via DCT for speech recognition
  \item Chroma features for music analysis
  \item Gammatone filters for perceptual accuracy
  \item Binaural spectrograms: ITD, IPD, ILD, ILR for spatial audio analysis
  \item MDCT/IMDCT with sine-window perfect reconstruction
  \item FFT convolution, deconvolution, and overlap-save streaming
  \item Minimum-phase conversion via the real-cepstrum method
\end{itemize}

\textbf{Optimizations:}
\begin{itemize}
  \item Plan-based computation
  \item Sparse matrix operations
  \item Workspace patterns (zero allocation in hot loops)
  \item Cache-efficient memory layouts
  \item Optional Rayon parallelisation (\code{rayon} feature)
\end{itemize}

\subsection{Applications and Extensions}

The techniques in this manual enable:

\begin{itemize}
  \item Speech recognition and speaker identification
  \item Music information retrieval
  \item Audio classification and tagging
  \item Sound event detection
  \item Acoustic scene analysis
  \item Image denoising and enhancement
  \item Texture analysis
  \item Pattern recognition
\end{itemize}

\textbf{Software:}
\begin{itemize}
  \item FFTW:\ \texttt{www.fftw.org}
  \item Librosa (Python): \texttt{librosa.org}
  \item Numpy (Python): \texttt{numpy.org}
  \item SciPy (Python): \texttt{scipy.org}
  \item This library: \texttt{github.com/jmg049/Spectrograms}
\end{itemize}
\clearpage
